---
title: 
csl: utils/plos.csl
fontsize: 10pt
output:
  bookdown::pdf_document2:
    includes:
      in_header: utils/header1_midatlantic.tex
    keep_tex: yes
    toc: false
    number_sections: false
  bookdown::html_document2: 
    df_print: paged
  bookdown::word_document2:
    toc: false
link-citations: yes
geometry: left=2cm, right=2cm, top=2cm, bottom=3cm, footskip = .5cm
subparagraph: yes
bibliography: utils/SOE2024.bib
urlcolor: blue
always_allow_html: true
params:
  region: "MidAtlantic"
  header: "utils/header1_midatlantic.tex"
---


```{r setup, include=FALSE}
# Default Rmd options
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  dev = "cairo_pdf",
  warning = FALSE,
  fig.width = 4,
  fig.asp = 0.45,
  fig.align = "center"
) # allows for inserting R code into captions

# Plotting and data libraries
library(tidyverse)
library(ecodata)
library(here)
library(flextable)
library(patchwork)

set_flextable_defaults(
  font.size = 9,
  padding.bottom = 0,
  padding.top = 1,
  line_spacing = 1,
  font.family = "Palatino"
)
```

\setcounter{page}{4}
\thispagestyle{fancy}

# Introduction

## About This Report

```{r, intro, results = "asis"}
if (params$region == "MidAtlantic") {
  res <- knitr::knit_child(
    text = "This report is for the Mid-Atlantic Fishery Management Council (MAFMC). The purpose of this report is to synthesize ecosystem information to allow the MAFMC to better meet fishery management objectives, and to update the MAFMC’s Ecosystem Approach to Fishery Management (EAFM) risk assessment. The major messages of the report are synthesized on pages 1 and 2, with highlights of 2024 ecosystem events on page 3.",
    quiet = TRUE
  )
} else if (params$region == "NewEngland") {
  res <- knitr::knit_child(
    text = "This report is for the New England Fishery Management Council (NEFMC). The purpose of this report is to synthesize ecosystem information to allow the NEFMC to better meet fishery management objectives. The major messages of the report are synthesized on pages 1-3, with highlights of 2024 ecosystem events on page 4.",
    quiet = TRUE
  )
}

cat(res, sep = "\n\n")
```
The information in this report is organized into two main sections; [performance measured against ecosystem-level management objectives](#performance-relative-to-fishery-management-objectives) (Table \ref{tab:management-objectives}), and potential [risks to meeting fishery management objectives](#risks-to-meeting-fishery-management-objectives) (Table \ref{tab:management-risks}: [climate change](#climate-and-ecosystem-change) and [other ocean uses](#other-ocean-uses-offshore-wind)). A final section highlights [notable 2024 ecosystem observations](#highlights).
