## Objective: Mid Atlantic Commercial Profits `r I(decreasing)` (revenue) `r I(no_trend)` (profitability) `r I(below_average)` (all)

:::: { .soe-50-50-slide}

::: {.soe-50-column}
Indicators: Commercial revenue

```{r}
#| out-height: "500px"
return_plot("comm-revenue", region = "MidAtlantic", mode = "slide")
```

On average, Mid Atlantic community revenue is derived species that are _highly sensitive_ to environmental variability, with no trend over time.

:::

::: {.soe-50-column}

New Indicator: Index of Revenue, Costs, and Profits from *federal trips only*

```{r}
return_plot("comdat-profit", region = "MidAtlantic", mode = "slide")
```

- Revenue shows a significant long-term decline, but profitability does not yet show a statistically significant decline.
- Trip costs do not include labor

:::

::::


## Objective: Mid Atlantic Commercial Profits `r I(no_trend)`  `r I(below_average)` 

::: { .soe-50-50-slide}

::: {.soe-50-column}

Indicator: Bennet--price and volume indices

```{r bennet}
# <!-- #| out-height: "350px" -->
return_plot("bennet", region = "MidAtlantic", mode = "slide")
```

- Recent low prices and volumes
- Especially in benthos (shellfish)

:::

::: {.soe-50-column}

```{r bennet-guild}
# <!-- #| out-height: "400px" -->
return_plot("bennet-all", region = "MidAtlantic", mode = "slide")
```

:::

::::