## Objective: New England Commercial Profits `r I(no_trend)`  `r I(below_average)` 

::: { .soe-50-50-slide}

::: {.soe-50-column}
Indicator: Commercial Revenue

```{r}
return_plot("comm-revenue", region = "NewEngland", mode = "slide")
```

Both regions' revenue depends on a single species sensitive to environmental changes (scallop and lobster)

Lower percent of total GOM revenue is coming from NEFMC-managed species

New England community revenue is _moderately dependent_ on species sensitive to environmental variability with no trend.

:::

::: {.soe-50-column}
Indicator: Bennet--price and volume indices

```{r}
#| out-height: "600px"
return_plot("bennet", region = "NewEngland", mode = "slide")
```

GB: Low 2023/2024 revenue driven by low volume of landings (benthos/shellfish)

GOM: Higher 2024 revenue driven by high prices (benthivores/lobster)

:::

::::

## Objective: New England Commercial Profits `r I(no_trend)`  `r I(below_average)` 

::: { .soe-50-50-slide}

::: {.soe-50-column}
Indicator: Profit Index

```{r}
return_plot("comdat-profit", region = "NewEngland", mode = "slide")
```

:::

::: {.soe-50-column}
<br>
<br>

New Indicator: Index of Revenue, Costs, and Profits from *federal trips only*

GB: High costs since 2021, but recent decrease helps offset low revenue

GOM: Profit generally follows revenue. 2024 showed average costs and profitability

:::

::::