## Objective: Mid Atlantic Ecological Stability: Not Stable

:::: {.soe-50-50-slide}

::: {.soe-50-column}

*Ecological* Indicators: Community Composition

```{r}
return_plot("exp-n", region = "MidAtlantic", mode = "slide")
```

```{r}
return_plot("zoo-diversity", region = "MidAtlantic", mode = "slide")
```

- Long-term changes in community composition 

:::

::: {.soe-50-column}

*Ecological* Indicators: Zooplankton biomass

```{r}
return_plot("zoopanom", region = "MidAtlantic", mode = "slide")
```

- Benthos biomass also shows trends

:::

::::

## Objective: Mid Atlantic Ecological Stability: Not Stable

*Ecological* Indicators: fish biomass & traits

:::: {.soe-50-50-slide}

::: {.soe-50-column}

```{r}
return_plot("nefsc-biomass-mab", region = "MidAtlantic", mode = "slide")
```

:::

::: {.soe-50-column}

```{r}
return_plot("traits-tl", region = "MidAtlantic")
```


- Declines in species trophic level & other life history traits suggests changing ecology
- Trends in biomass would not be expected in a stable system
- Productivity of managed species is lower than historic levels

:::

::::

