## Objective: Mid Atlantic Ecological Stability: Not Stable

*Ecological* Indicators: Community Composition

:::: {.soe-50-50-slide}

::: {.soe-50-column}

```{r}
return_plot("exp-n", region = "MidAtlantic", mode = "slide")
```

- Long-term changes in community composition 

:::

::: {.soe-50-column}

```{r}
return_plot("zoo-community", region = "MidAtlantic", mode = "slide")
```

```{r}
return_plot("zoo-diversity", region = "MidAtlantic", mode = "slide")
```

:::

::::

## Objective: Mid Atlantic Ecological Stability: Not Stable

*Ecological* Indicators: fish biomass

:::: {.soe-50-50-slide}

::: {.soe-50-column}

```{r}
return_plot("nefsc-biomass-mab", region = "MidAtlantic", mode = "slide")
```

:::

::: {.soe-50-column}

```{r}
return_plot("foragebio", region = "MidAtlantic", mode = "slide")
```

- Trends in biomass would not be expected in a stable system

:::

::::

## Objective: Mid Atlantic Ecological Stability: Not Stable

*Ecological* Indicators: benthos and zooplankton biomass

:::: {.soe-50-50-slide}

::: {.soe-50-column}

```{r}
return_plot("benthos", region = "MidAtlantic", mode = "slide")
```
:::

::: {.soe-50-column}

```{r}
return_plot("zoopanom", region = "MidAtlantic", mode = "slide")
```

:::

::::

## Objective: Mid Atlantic Ecological Stability: Not Stable

*Ecological* Indicators: fish traits

:::: {.soe-50-50-slide}

::: {.soe-50-column}

```{r}
return_plot("traits-k", region = "MidAtlantic")
```

```{r}
return_plot("traits", region = "MidAtlantic")
```

:::

::: {.soe-50-column}

```{r}
return_plot("traits-tl", region = "MidAtlantic")
```


* Declines in species trophic level & other life history traits suggests changing ecology

:::

::::

## Objective: Mid Atlantic Ecological Stability: Not Stable

:::: {.soe-50-50-slide}

::: {.soe-50-column}

*Ecological* Indicators: fish productivity

- Declines in productivity of managed species
- But no trend in aggregate biomass at the guild level suggests the system is able to adapt to change

:::

::: {.soe-50-column}

```{r}
return_plot("productivity-anomaly", region = "MidAtlantic", mode = "slide")
```

:::

::::