## Objective: Georges Bank Ecological Stability: Not Stable

:::: {.soe-50-50-slide}

::: {.soe-50-column}

*Ecological* Indicators: zooplankton communities

```{r}
return_plot("zoo-community", region = "NewEngland", mode = "slide")
```

* GB seeing large-scale shifts in the broad species that make up zooplankton communities

:::

::: {.soe-50-column}

*Ecological* Indicators: Survey Biomass

```{r}
#| fig-height: 3.5
#| fig-width: 6
#| out-height: "600px"
return_plot("nefsc-biomass-gb", region = "NewEngland", mode = "slide")
```

* Biomass of large functional groups is changing for some but not all species suggesting a change in fish communities


:::

::::

## Objective: Gulf of Maine Ecological Stability: Not Stable

:::: {.soe-50-50-slide}

::: {.soe-50-column}

*Ecological* Indicators: fish traits

```{r}
return_plot("traits-tl", region = "NewEngland", mode = "slide")
```

* Finfish with a higher trophic level in spring
* Finfish show a shorter length at maturity, lower fecundity, and faster growth rates. 

:::

::: {.soe-50-column}

*Ecological* Indicators: Survey Diversity

```{r}
return_plot("exp-n", region = "NewEngland", mode = "slide")
```

* Increase in number of species caught in trawl survey
* Fish communities are experience large-scale changes potentially chaning structure and function of ecosystem

:::

::::