## Objectives: Coastwide Protected species *Maintain bycatch below thresholds* `r I(mixed_trend)`  `r I(meeting_objectives)` 

:::: {.soe-50-50-slide}

::: {.soe-50-column}

Indicators: Harbor porpoise and gray seal bycatch

```{r}
return_plot("harborporpoise", region = "BothReports")
```

```{r,  fig.height=3}
return_plot("grayseal", region = "BothReports")
```

:::

::: {.soe-50-column}

Implications: 

* High uncertainty in gray seal bycatch estimates but considered meeting objectives
* Gray seal are among the highest bycatch of any U.S. marine mammal. The increasing trend in gray seal bycatch may be related to an increase in the gray seal population (U.S. pup counts).
```{r, eval = FALSE, results = "asis"}
cat('* Risk element: TechInteract, evaluated by species and sector:  14 <span style="background-color:lightgreen;">low</span>, 7 <span style="background-color:yellow;">low-mod</span>, 2 <span style="background-color:orange;">mod-high</span> risk')
```

* The downward trend in harbor porpoise bycatch can also be due to a decrease in harbor porpoise abundance in US waters, reducing their overlap with fisheries, and a decrease in gillnet effort. 

:::

::::

## Objectives: Coastwide Protected species *Recover endangered populations* `r I(mixed_trend)` `r I(below_average)`

:::: {.soe-50-50-slide}

::: {.soe-50-column}

Indicators: North Atlantic right whale population, calf counts

```{r}
return_plot("narw-abundance", region = "BothReports")
```

```{r}
return_plot("NARW-calf-abundance", region = "BothReports")
```


:::

::: {.soe-50-column}

Implications:  

* No change in adult populations since 2020 but mortality still above recovery rate
* Interactions between North Atlantic Right Whales (NARW) and fixed gear fisheries and offshore wind development
* Distribution shifts  driven by copepod availability. 
* Unusual mortality events continue for 3 large whale species.
* Gray seal population is increasing

```{r salmon, eval = params$region == "NewEngland", results = "asis"}
res <- knitr::knit_child(text = '* Atlantic salmon return rates continue to decline
\n\n
`r return_plot("salmon", region = "NewEngland")`',
quiet = TRUE)

cat(res, sep = "\n\n")
```

:::

::::