## Objective: New England Recreational opportunities: Effort `r I(no_trend)` `r I(below_average)`

:::: {.soe-50-50-slide}

::: {.soe-50-column}

Indicators: Recreational effort

```{r}
return_plot("rec-op", region = "NewEngland", mode = "slide")
```


:::

::: {.soe-50-column}

Implications

* Recreational effort peaked in 2010 but is now stable

* Lack of a long-term trend suggests a consistent amount of recreational opportunities in New England

:::

::::

## Objective: New England Recreational opportunities: Fleet and Catch Diversity `r I(mixed_trend)` `r I(near_average)`

:::: {.soe-50-50-slide}

::: {.soe-50-column}

Indicators: Recreational catch and fleet diversity  

```{r }
return_plot("rec-div", region = "NewEngland", mode = "slide")
```

```{r }
return_plot("recdat-div-catch", region = "NewEngland", mode = "slide")
```

:::

::: {.soe-50-column}

Implications

* No long-term change in types of recreational trips
* *But* there is an increase in diversity of species caught recreationally
* This suggests recreational fishers are catching a wider variety of species

:::

::::