## Risks: Offshore Wind Development New England

:::: {.soe-50-50-slide}

::: {.soe-50-column}

Indicators: fishery and community specific revenue in lease areas

```{r}
return_plot("wea-spp-rev", region = "NewEngland", mode = "slide")
```

- Although sea scallop has the highest revenue in Active Project areas, representing $\leq 1$% of the total revenue for the fishery. 
- Some groundfish and skates have up to 13% of their revenue from Active Project areas, but these species have lower overall revenue than scallops.
<!-- - Wind lease areas that are under construction or complete generally represent <5% of port-level revenue for New England ports, with the exceptions of Little Compton, RI and Westport, MA. -->
<!-- - Wind lease areas that have not yet begun construction generally represent <5% of port-level revenue for New England ports, with the exception of Boston, MA. -->

:::

::: {.soe-50-column}

```{r}
return_plot("wea-port-rev", region = "NewEngland", mode = "slide")
```

:::

::::

## Risks: Offshore Wind Development New England

:::: {.soe-50-50-slide}

::: {.soe-50-column}

- Council request to look at NEFMC-managed species that are being caught in the wind lease areas and landed in the Mid-Atlantic
- Point Lookout, NY is the only Mid-Atlantic port landing >5% of its revenue from NEFMC-managed species in the wind lease areas

:::

::: {.soe-50-column}

```{r}
return_plot("wind-rev", region = "NewEngland", mode = "slide")
```

:::

::::