## Risks to Managing Seasonally: Mid-Atlantic

:::: {.soe-50-50-slide}

::: {.soe-50-column}

Drivers: thermal transition, habitat persistence, bloom timing

```{r, fig.alt="Ocean summer length: the annual total number of days between the spring thermal transition date and the fall thermal transition date (black), with an increasing trend (orange)."}
return_plot("transition", region = "MidAtlantic", mode = "slide")
```

Cold pool seasonal persistence

```{r, fig.alt="Cold pool persistence index based on bias-corrected ROMS-NWA (open circles) and GLORYS (closed circles)."}
return_plot("cold-pool-time", region = "BothReports", mode = "slide")
```

:::

::: {.soe-50-column}

Bloom timing

```{r, fig.alt="Monthly median chlorophyll a concentration in the MAB (black) with significant increase in January (orange line) and decrease in September (purple line)."}
return_plot("chl-month", region = "MidAtlantic", mode = "slide")
```

Future considerations 

*  Management actions that rely on effective alignment of fisheries availability and biological processes should continue to evaluate whether prior assumptions on seasonal timings still hold.

*  New indicators should be developed to monitor timing shifts for stocks.

:::

::::
