## Risks to Setting Catch Limits: New England Indicators

:::: {.soe-50-50-slide}

::: {.soe-50-column}

Fish productivity

```{r productivity-anomaly}
#| out-height: "750px"
return_plot("productivity-anomaly", region = "NewEngland", mode = "slide")
```

:::

::: {.soe-50-column}

**Management Concerns**

* Sustainable management and rebuilding plans rely on assumptions about fish productivity (growth, reproduction, and natural mortality)
* Even if not understood well, environmental changes can impact biological processes and reference points

**Fish Productivity**

* Productivity is the net result of growth, reproduction, and natural mortality
* Fish productivity had been low since 2020

:::

::::

## Risks to Setting Catch Limits: New England Indicators
Fish Condition

```{r}
#| out-height: "650px"
#| fig-align: center
return_plot("ne-cf", region = "NewEngland", mode = "slide")
```

Condition relates to reproduction potential and market quality. High variability but patterns emerge across species.

## Risks to Setting Catch Limits: New England Drivers

:::: {.soe-50-50-slide}

::: {.soe-50-column}

Forage Quality and Abundance

```{r foragebio}
return_plot("foragebio", region = "NewEngland", mode = "slide")
```

Forage fish abundance/quality and benthic prey changes annually.
Declines in spring macrobenthos may impact juvenile fish

:::

::: {.soe-50-column}

benthos abundance

```{r benthos}
#| out-height: "700px"
return_plot("benthos", region = "NewEngland", mode = "slide")
```

:::

::::

## Risks to Setting Catch Limits: New England Drivers

:::: {.soe-50-50-slide}

::: {.soe-50-column}

Low trophic level processes

```{r totppagain}
return_plot("totpp", region = "NewEngland", mode = "slide")
```

* Primary production determines the energy available to the ecosystem
* Forage and juvenile fish and marine mammals rely on zooplankton to grow. Shifts in zooplankton communities can change the overall quality of food. 

:::

::: {.soe-50-column}

```{r zoopanom}
#| out-height: "800px"
return_plot("zoopanom", region = "NewEngland", mode = "slide")
```

:::

::::

## Risks to Setting Catch Limits: New England Drivers

Changes in fish traits

Finfish communities are changing in GOM towards species with higher trophic levels and faster growth rates

:::: {.soe-50-50-slide}

::: {.soe-50-column}

```{r}
return_plot("traits-k", region = "NewEngland", mode = "slide")
```

:::

::: {.soe-50-column}

```{r}
return_plot("traits-tl", region = "NewEngland", mode = 'slide')
```

:::

::::

## Risks to Setting Catch Limits: New England Drivers

Environmental: Habitat Quality and Circulation

```{r, eval = TRUE}
#| fig-align: center
#| out-height: "350px"
return_plot("mab-oa", region = "BothReports",mode = 'slide')
```


:::: {.soe-50-50-slide}

::: {.soe-50-column}

```{r, eval = TRUE}
#| out-height: "350px"
return_plot("advection-index", region = 'NewEngland', mode = 'slide')
```


:::

::: {.soe-50-column}

* Episodic habitat risk events for scallops and squid present but sparse in 2025
* New Indicator: Advection off shelf (how much are larvae getting carried offshore?)
  * Mid and Deep water more likely to move onshore in spring over time

:::

::::
