
## Revised Risks: Climate and Ecosystem Change 

:::: {.soe-50-50-slide}

::: {.soe-50-column}

<br> <br>

```{r risktabagain}
mng_obj <- data.frame("Risks" = c("Risks to Managing Spatially",
                                  "Risks to Managing Seasonally",
                                  "Risks to Setting Catch Limits",
                                  "Offshore Wind Risks"),
                      "Observation Indicators" = c("Managed species (fish and cetacean) distribution shifts",
                                                   "Managed species spawning and migration timing changes", 
                                                   "Managed species body condition and recruitment changes",
                                                   "Fishery revenue and landings from wind lease areas by species and port"),
                      "Potential Driver Indicators" = c("Benthic and pelagic forage distribution; ocean temperature, changes in currents and cold pool",
                                                        "Habitat timing: Length of ocean summer, cold pool seasonal persistence",
                                                        "Benthic and pelagic forage quality & abundance: ocean temperature & acidification ",
                                                        "Wind development map; Protected species presence and  hotspots"
                      ))

mng_obj$service <- c(rep("Climate and Ecosystem Risks", 3), 
                     rep("Other Ocean Uses Risks", 1))

tab <- flextable::as_grouped_data(mng_obj, groups = "service") %>% 
  flextable::as_flextable(hide_grouplabel=TRUE) %>% 
  flextable::align(i = ~ !is.na(service), align = "left") %>% 
  flextable::bold(i = ~ !is.na(service), bold = TRUE) %>%
  flextable::set_header_labels(Risks = "Risk categories",
                               Observation.Indicators = "Observation indicators reported",
                               Potential.Driver.Indicators = "Potential driver indicators reported") %>%

  flextable::set_caption("Risks to meeting fishery management objectives in the Mid-Atlantic Bight")  |>
flextable::fontsize(size = 16,
part = "all")
  #flextable::autofit() 
```

```{r}
FitFlextableToPage(tab)
```

:::

::: {.soe-50-column}

### Risks to Managing Spatially
Potential Impacts: Spatial mis-allocation of quotas may lead to unmet quotas, increased discards, and/or miscalculated fishing targets. 
### Risks to Managing Seasonally
Potential Impacts: Spawning closures, seasonal openings, and seasonal quota allocations may be less effective if mis-timed with biological events, resulting in decreased seafood production.

### Risks to Setting Catch Limits
Potential Impacts: Unaccounted for and unknown productivity changes may lead to mis-specified quotas and rebuilding plans, especially if they are not considered in stock reference points and short-term stock projections.

:::

::::

## Risks to Managing Spatially: Indicators

:::: {.soe-50-50-slide}

::: {.soe-50-column}

Fish distribution shifts

```{r}
return_plot("species-dist", region = "BothReports", mode = "slide")
```  

Significant species distributions seen for fish, HMS, and mammals
:::

::: {.soe-50-column}

Cetacean distribution shifts 

```{r}
return_plot("protectedspp-dist-shifts", region = "BothReports", mode = "slide")
```

Changing distributions can impact:

  * spatial allocation & cross-jurisdictional management
  * effectiveness of bycatch reduction measures
  * Availability to surveys and fisheries

:::

::::

## Risks to Managing Spatially: Biological Drivers

:::: {.soe-50-50-slide}

::: {.soe-50-column}

Forage fish distributions

```{r forageshifts, fig.alt="Eastward (left) and northward (right) shifts in the center of gravity for 20 forage fish species on the Northeast U.S. Shelf, with increasing trend (orange) for fall eastward and northward center of gravity."}
return_plot("forageshifts", region = "BothReports", mode = "slide")
```

Some species are following their food, but not all.
Both pelagic and benthic prey are shifting towards the northeast

:::

::: {.soe-50-column}

Macrobenthos (mid-size benthic species) distributions

```{r macrobenthosshifts, fig.alt="Eastward (left) and northward (right) shifts in the center of gravity for macrobenthos species on the Northeast U.S. Shelf"}
return_plot("macrobenthosshifts", region = "BothReports", mode = "slide")
```

Not all species are as mobile as their prey. 
Can cause a mismatch between fish and their prey

:::

::::

## Risks to Managing Spatially: Biological Drivers

Shifts in zooplankton (from ECOMON)

:::: {.soe-50-50-slide}

::: {.soe-50-column}

```{r}
return_plot("calfin-cog", region = "NewEngland", mode = "slide")
```

_Calanus_ shifting westward in last decade

:::

::: {.soe-50-column}


```{r}
return_plot("smallcopeall-cog", region = "BothReports", mode = "slide")
```

Long-term northeast shifts in Small copepods

:::

::::

## Risks to managing spatially: Environmental Drivers

Thermal Habitat Changes 

:::: {.soe-50-50-slide}

::: {.soe-50-column}

```{r long-term-sst, fig.alt="Northeast US annual sea surface temperature (SST, black), with increasing trend (orange)."}
return_plot("long-term-sst", region = "BothReports", mode = "slide")
```

```{r GSI, fig.alt = "Index representing changes in the location of the Gulf Stream north wall (black). Positive values represent a more northerly Gulf Stream position, NO LONGER HAS increasing trend."}
return_plot("GSI", region = "BothReports", mode = "slide")
```

* The entire shelf is getting warmer with northward shifts in Gulf Stream
* Since 2023 - sharp drop in gulf stream and temperature
* Changes in currents and temperature can alter habitat suitability for species
* Mobile species can follow changes, but others cannot (i.e. scallops)
:::

::: {.soe-50-column}

```{r cold-pool-size, fig.alt="Seasonal cold pool mean temperature (left) and spatial extent index (right), based on bias-corrected ROMS-NWA (open circles) and GLORYS (closed circles), with declining trends (purple)."}
#| out-height: "500px"
return_plot("cold-pool-size", region = "BothReports", mode = "slide")
```


* Some changes can be local and variable
  * Mid-Atlantic Cold pool important regional feature that is shrinking and warming habitat
  * Last 3 years saw an improvement in cold pool conditions


:::

::::