## Risks to Managing Seasonally: New England

:::: {.soe-50-50-slide}

::: {.soe-50-column}

Drivers: thermal transition, habitat persistence, bloom timing

```{r transition}
return_plot("transition", region = "NewEngland", mode = "slide")
```

:::

::: {.soe-50-column}

Future considerations 

*  Management actions that rely on effective alignment of fisheries availability and biological processes should continue to evaluate whether prior assumptions on seasonal timings still hold.

*  New indicators should be developed to monitor timing shifts for stocks.

:::

::::

Bloom timing

```{r montlychl}
return_plot("chl-month", region = "NewEngland", mode = "slide")
```

## Risks to Setting Catch Limits: New England

Indicator: Fish productivity

```{r productivity-anomaly}
return_plot("productivity-anomaly", region = "NewEngland", mode = "slide")
```

## Risks to Setting Catch Limits: New England

Indicator: Condition

```{r}
return_plot("ne-cf", region = "NewEngland", mode = "slide")
```

## Risks to Setting Catch Limits: New England Drivers

:::: {.soe-50-50-slide}

::: {.soe-50-column}

Drivers: Forage Quality and Abundance

```{r foragebio}
return_plot("foragebio", region = "NewEngland", mode = "slide")
```

:::

::: {.soe-50-column}

New indicators: benthos abundance

```{r benthos}
return_plot("benthos", region = "NewEngland", mode = "slide")
```

:::

::::

## Risks to Setting Catch Limits: New England Drivers

:::: {.soe-50-50-slide}

::: {.soe-50-column}

Drivers: Low trophic levels

```{r totppagain}
return_plot("totpp", region = "NewEngland", mode = "slide")
```

:::

::: {.soe-50-column}

```{r zoopanom}
return_plot("zoopanom", region = "NewEngland", mode = "slide")
```

:::

::::

## Risks to Setting Catch Limits: Coastwide

Drivers: Environmental   

*2024 Thermal habitat area by depth*

```{r}
b <- ecodata::plot_thermal_habitat_area(report = "NewEngland", EPU="GB") + ggplot2::theme(legend.position = "none")
c <- ecodata::plot_thermal_habitat_area(report = "NewEngland", EPU="GOM") + ggplot2::theme(legend.position = "bottom")

ggpubr::ggarrange(b, c, nrow = 1)
```
