## Risks to Managing Seasonally: Indicators

Spawning timing & changes in migration patterns

:::: {.soe-50-50-slide}

::: {.soe-50-column}

```{r spawntiming, fig.alt="Percent resting stage (non-spawning) mature female fish (black) with significant increases (orange) and decreases (purple) from two haddock and three yellowtail flounder stocks: CC = Cape Cod Gulf of Maine, GOM = Gulf of Maine, GB = Georges Bank, SNE = Southern New England."}
return_plot("spawntiming", region = "BothReports", mode = "slide")
```

```{r, eval = params$region != "MidAtlantic"}
return_plot("cold-pool-time", region = "BothReports", mode = "slide")
```


:::

::: {.soe-50-column}

**Life Cycle Events**

*  Effective fishing seasons & openings/closing rely on alignment with seasonal events in fish stocks
*  Though limited, evidence of large-scale changes in spawn timing

**Migratory Species**

*  Recreational bigeye tuna caught [50 days earlier](https://noaa-edab.github.io/catalog/timing_shifts.html#key-results-and-visualizations-59) and bluefin tuna 38-80 days earlier (2002 to 2019).
*  Right and humpback whale peak spring habitat has shifted 18-19 days later.
*  Baseline information on [large whale seasonal presence](https://noaa-edab.github.io/catalog/cetacean_acoustic.html) has been collected.

:::

::::
