## Objective: Mid Atlantic Seafood production `r I(decreasing)` (total and recreational) `r I(no_trend)` (managed) `r I(below_average)` 

:::: {.soe-50-50-slide}

::: {.soe-50-column}
Indicators: Commercial landings, climate vulnerability

```{r}
#| out-height: "500px"
return_plot("total-landings", region = "MidAtlantic", mode = "slide")
```

Multiple potential drivers: ecosystem and stock production, management, market conditions, and environmental change.

Mid Atlantic community landings are _moderately dependent_ on species sensitive to environmental variability, with no trend over time.

:::

::: {.soe-50-column}

Indicators: Recreational harvest

```{r}
#| out-height: "250px"
return_plot("rec-landings", region = "MidAtlantic", mode = "slide")
```

```{r}
#| out-height: "400px"
return_plot("rec-hms", region = "MidAtlantic", mode = "slide")
```

Drivers: Shifts in fishing behavior (landed vs catch-and-release) and closure of mako fishery

:::

::::

## Mid Atlantic Landings drivers: Stock status? TAC? 

:::: {.soe-50-50-slide}

::: {.soe-50-column}

Indicator: Stock status

```{r stock-status}
#| out-height: "500px"
return_plot("stock-status", region = "MidAtlantic", mode = "slide")
```

```{r abcacl}
#| out-height: "250px"
return_plot("abcacl-catch", region = "MidAtlantic", mode = "slide")
```
:::

::: {.soe-50-column}

Indicators: Total ABC or ACL, and realized catch relative to management target

```{r}
# <!-- #| out-height: "650px" -->
return_plot("abcacl-stacked", region = "MidAtlantic", mode = "slide")

```

- Two groups of species:
  - Catch ratios near 1
  - Catch ratios near 0.3
- Species with catch ratios near 1 may be constrained by total allowable catch; although most stock biomass is meeting management targets, the amount of catch supported may be lower than historic 
- Species with lower catch ratios may be constrained by spatial, temporal, or other non-quota-related management factors

:::

::::

## Implications: Mid Atlantic Seafood Production Drivers

::::{.soe-50-50-slide}

::: {.soe-50-column}

Regional biomass does not appear to drive landings trends

```{r}
#| out-height: "650"
return_plot("nefsc-biomass-mab", region = "MidAtlantic", mode = "slide")
```

Recreational drivers differ: shark fishery management, possibly survey methodology

:::

::: {.soe-50-column}

Declining aggregate planktivores, benthos?

```{r}
#| out-height: "500px"
return_plot("comm-landings", region = "MidAtlantic", mode = "slide")
```

Monitor: 

* Climate risks including warming, ocean acidification, and shifting distributions  

* Ecosystem composition and production changes  

* Fishing engagement  

:::

::::
