## Objective: Mid Atlantic Seafood production `r I(decreasing)` `r I(below_average)` 

:::: {.soe-50-50-slide}

::: {.soe-50-column}
Indicators: Commercial landings, climate risk

```{r}
#| out-height: "500px"
return_plot("total-landings", region = "MidAtlantic", mode = "slide")
```

```{r climatevul-land, fig.alt="Mid-Atlantic region total climate vulnerability of commercial landings (sum of Mid-Atlantic port landings weighted by species climate vulnerability from Hare et al. 2016)."}
return_plot("climatevul-land", region = "MidAtlantic", mode = "slide")
```
:::

::: {.soe-50-column}

Indicators: Recreational harvest

```{r}
#| out-height: "250px"
return_plot("rec-landings", region = "MidAtlantic", mode = "slide")
```

```{r}
#| out-height: "400px"
return_plot("rec-hms", region = "MidAtlantic", mode = "slide")
```

Multiple potential drivers: ecosystem and stock production, management, market conditions, and environmental change.

:::

::::

## Mid Atlantic Landings drivers: Stock status? TAC? 

:::: {.soe-50-50-slide}

::: {.soe-50-column}

Indicator: Stock status

```{r stock-status}
#| out-height: "500px"
return_plot("stock-status", region = "MidAtlantic", mode = "slide")
```

```{r abcacl}
#| out-height: "250px"
return_plot("abcacl-catch", region = "MidAtlantic", mode = "slide")
```
:::

::: {.soe-50-column}

Indicators: Total ABC or ACL, and Realized catch relative to management target

```{r}
#| out-height: "650px"
return_plot("abcacl-stacked", region = "MidAtlantic", mode = "slide")

```

Catch near limits on average

:::

::::

## Implications: Mid Atlantic Seafood Production Drivers

::::{.soe-50-50-slide}

::: {.soe-50-column}

Biomass does not appear to drive landings trends

```{r}
#| out-height: "650"
return_plot("nefsc-biomass-mab", region = "MidAtlantic", mode = "slide")
```


Recreational drivers differ: shark fishery management, possibly survey methodology

:::

::: {.soe-50-column}

Declining aggregate planktivores, benthos?

```{r}
#| out-height: "500px"
return_plot("comm-landings", region = "MidAtlantic", mode = "slide")
```

Monitor: 

* climate risks including warming, ocean acidification, and shifting distributions  
* ecosystem composition and production changes  
* fishing engagement  

:::

::::
