## Objective: New England Seafood production `r I(decreasing)` `r I(no_trend)` `r I(below_average)`

:::: {.soe-50-50-slide}

::: {.soe-50-column}

Indicators: Commercial landings, climate risk

```{r}
return_plot("total-landings", region = "NewEngland", mode = "slide")
```

Drivers: ecosystem and stock production, management actions (stock rebuilding), market conditions, and environmental change

New England community landings are _moderately dependent_ on species sensitive to environmental variability, but this risk has been increasing.

:::

::: {.soe-50-column}
Indicators: Recreational harvest

```{r}
return_plot("rec-landings", region = "NewEngland", mode = "slide")
```

```{r}
return_plot("rec-hms", region = "NewEngland", mode = "slide")
```
Drivers: Shifts in fishing behavior (landed vs catch-and-release) and closure of mako fishery

:::

::::



## New England drivers: Stock status?
:::: {.soe-50-50-slide}

::: {.soe-50-column}
Indicator: Stock status

```{r}
return_plot("stock-status", region = "NewEngland", mode = "slide")
```

:::

::: {.soe-50-column}

Changes: 

* B/Bmsy > 1 GB Haddock & Plaice 
* F/Fmsy < 1 GOM Haddock

All Cod stocks have B/Bmsy < 1

17 stocks below target, 10 below thresholds, and 12 with unknown status

_Stock status and required management actions still likely playing large role in seafood declines._

:::

::::

## New England drivers: Survey biomass
:::: {.soe-50-50-slide}

::: {.soe-50-column}
Indicator: Survey Biomass

```{r}
#| fig-cap: "Gulf of Maine"
#| fig-height: 3.5
#| fig-width: 6
#| out-height: "600px"
return_plot("nefsc-biomass-gom", region = "NewEngland", mode = "slide")
```

Increasing guild biomass in planktivores & benthos *despite* declining landings and
Piscivores more present in fall than spring survey

:::

::: {.soe-50-column}

&nbsp;

```{r}
#| fig-cap: "Georges Bank"
#| fig-height: 3.5
#| fig-width: 6
#| out-height: "600px"
return_plot("nefsc-biomass-gb", region = "NewEngland", mode = "slide")
```

System is *productive overall* just not for targeted species

:::

::::

## Implications: New England Seafood Production 

::::{.soe-50-50-slide}

::: {.soe-50-column}

Drivers:

* Decline in commercial landings is most likely driven by market dynamics and actions needed to rebuild individual stocks (lower quotas) 
* Drivers of Recreational landings differ: tighter shark fishery regulations, changing demographics and preferences of anglers

Monitor: 

* Climate & environmental risks: shifting distributions, changes to seasonal timing, and changes in system productivity  
* Changes to ecosystem composition and stability 
* Fishing engagement and behavior

:::

::: {.soe-50-img}

```{r}
return_plot("comm-landings", region = "NewEngland", mode = "slide")
```

:::

::::
