## Objective: New England Seafood production `r I(decreasing)` `r I(no_trend)` `r I(below_average)`

:::: {.soe-50-50-slide}

::: {.soe-50-column}
Indicators: Commercial landings, climate risk

```{r}
return_plot("total-landings", region = "NewEngland", mode = "slide")
```

:::

::: {.soe-50-column}
Indicators: Recreational harvest

```{r}
return_plot("rec-landings", region = "NewEngland", mode = "slide")
```

```{r}
return_plot("rec-hms", region = "NewEngland", mode = "slide")
```

:::

::::

Multiple drivers: ecosystem and stock production, management actions (stock rebuilding), market conditions (including COVID-19 disruptions), and environmental change

## New England drivers: Stock status? Survey biomass?
:::: {.soe-50-50-slide}

::: {.soe-50-column}
Indicator: Stock status

```{r}
return_plot("stock-status", region = "NewEngland", mode = "slide")
```

One more stock below BMSY from last year (S Silver Hake). No change in stocsk below 1/2 BMSY. Stock status and required management actions still likely playing large role in seafood declines.

:::

::: {.soe-50-column}

Indicators: Survey biomass

::: {layout-ncol=2}

```{r}
return_plot("nefsc-biomass-gom", region = "NewEngland", mode = "slide")
```

```{r}
return_plot("nefsc-biomass-gb", region = "NewEngland", mode = "slide")
```

:::
Biomass availability still seems unlikely driver

:::

::::

## Implications: New England Seafood Production 

::::{.soe-50-50-slide}

::: {.soe-50-column}

Drivers:

* Decline in commercial landings is most likely driven by actions to rebuild individual stocks (lower quotas) as well as market dynamics
* other drivers affecting recreational landings: tighter shark fishery regulations, changing demographics and preferences of anglers

Monitor: 

* Climate risks including warming, ocean acidification, and shifting distributions  
* Ecosystem composition and production changes  
* Fishing engagement

```{r}
return_plot("species-dist", region = "BothReports", mode = "slide")
```

:::

::: {.soe-50-img}

```{r}
return_plot("comm-landings", region = "NewEngland", mode = "slide")
```

:::

::::
