## Take Home

:::: {.soe-50-50-slide}

::: {.soe-50-column}

**Mostly meeting individual stock objectives.**

```{r}
return_plot("stock-status", region = "MidAtlantic")
```

Species distributions, timing, and productivity have changed. Ocean conditions highly variable year to year.

:::

::: {.soe-50-column}

**Ecosystem level management objectives worrying?**

Declining <span style="background-color:orange;">seafood production</span> 

Historically low <span style="background-color:orange;">revenue</span> 

Declining range of <span style="background-color:red;">recreational opportunities</span> 

Most community level revenue has high climate vulnerability

```{r}
return_plot("commvulex", region = "MidAtlantic")
```

:::

::::

## Discussion

:::: {.soe-50-50-slide}

::: {.soe-50-column}

<br>

```{r management-objectives-review}
mng_obj <- data.frame("Objective Categories" = c("Seafood Production",
                                                 "Commercial Profits","Recreational Opportunities",
                                                 "Stability","Social & Cultural",
                                                 "Protected Species",
                                                 "Management","Biomass", "Environment"
                                                 #,"Productivity",
                                                 #"Trophic structure","Habitat"
                                                 ),
"Indicators reported" = c("Landings; commercial total and by feeding guild; recreational harvest",
                               "Revenue decomposed to price and volume",
                               "Angler trips; recreational fleet diversity",
                               "Diversity indices (fishery and ecosystem)",
                               "Community fishing engagement and social vulnerability status",
                               "Bycatch; population (adult and juvenile) numbers; mortalities",
                               "Stock status; catch compared with catch limits",
                               "Biomass or abundance by feeding guild from surveys",
                               "Climate and ecosystem risk indicators listed in Table 2"
                          ))



mng_obj$service <- c(rep("Objectives: Provisioning and Cultural Services", 6), 
                     rep("Potential Drivers: Supporting and Regulating Services", 3))

mng_obj <- as.tibble(mng_obj)

flextable::as_grouped_data(mng_obj, groups = "service") %>% 
  flextable::as_flextable(hide_grouplabel=TRUE) %>% 
  flextable::align(i = ~ !is.na(service), align = "left") %>% 
  flextable::bold(i = ~ !is.na(service), bold = TRUE) %>%
  flextable::set_header_labels(Objective.Categories = "Objective categories",
                               Indicators.reported = "Indicators reported") %>%
  flextable::set_caption("Ecosystem-scale fishery management objectives") |>
flextable::fontsize(size = 16,
part = "all") %>%
  flextable::autofit() 
```

:::

::: {.soe-50-column}

**Review objectives: maximum benefit to the Nation in first 3 categories**

*  Is the list complete? 

*  Does the Council have reference points, desired states, or other performance metrics we should address for each objective?

Stability objective: are current indicators addressing Council needs?

Forecasts: what decisions or contexts would ocean forecasts be most useful for?

:::

::::