
```{r, eval = params$id_child_docs, results = "asis"}
<<child-header>>
```

```{r ne-draft, eval = params$draft}
# reset conditional values
<<conditional-values>>
```

### Indicators: Landings; commercial and recreational

This year, we present updated indicators for total [commercial landings](https://noaa-edab.github.io/catalog/comdat.html) (includes seafood, bait, and industrial landings), U.S. seafood landings (excludes industrial and bait uses) , and Council-managed U.S. seafood landings through 2024. Results show there are long-term declines in all New England landings time series (Fig. \ref{fig:total-landings}).
 
```{r, total-landings, fig.width = 6.5, fig.asp = 0.4, fig.cap = return_caption(chunk_name = "total-landings", region = "NewEngland")}
return_plot("total-landings", region = "NewEngland")
```

Commercial landings by guild include all species and all uses, and are reported as total for the guild and the NEFMC managed species within the  [guild](https://noaa-edab.github.io/catalog/aggregate_biomass.html). As reported in previous years, downward trends persist for piscivores and benthivores in both regions. Current high total landings for benthivores (GOM) are attributable to American lobster, and high benthos landings (GB) is attributable to clams and scallops (Fig. \ref{fig:comm-landings}). Current landings of planktivores and piscivores are still below the long-term mean.

Not all [aquaculture production](https://noaa-edab.github.io/catalog/aquaculture.html)is included in total seafood landings. In 2022, the Northeast region produced approximately 6,300 metric tons of aquacultured shellfish, with revenue of $133 million (Fisheries of the United States, 2022). 

```{r, comm-landings, fig.cap = return_caption(chunk_name = "comm-landings", region = "NewEngland"), fig.width=6.5, fig.asp=1}
return_plot("comm-landings", region = "NewEngland")
```

[Total Community Climate Change Risk](https://noaa-edab.github.io/catalog/community_climate_vulnerability.html) indicators evaluate port level risk to environmental variability based on dependence on species and their respective bioenvironmental vulnerabilities as assessed by regional experts. Total Vulnerability measures how much of a region’s landings (or revenue) is dependent on species sensitive to different climate and environmental change factors including temperature and acidification. For New England ports, the Total Vulnerability of landings (Fig. \ref{fig:climatevul-land}) was moderate in 2024 with no suggestion of a long-term trend.

```{r, climatevul-land, fig.cap = return_caption(chunk_name = "climatevul-land", region = "NewEngland"), fig.width=6.5, fig.asp=0.5}
return_plot("climatevul-land", region = "NewEngland")
```

Overall, [recreational harvest](https://noaa-edab.github.io/catalog/recdat.html) (retained fish presumed to be eaten) has declined in New England (Fig. \ref{fig:rec-landings}). Recent harvest remains near a time series low. The recent low in pelagic shark landings is largely driven by regulatory changes implemented in 2018, followed by the closure of the shortfin mako fishery in 2022. These actions were intended to rebuild the North Atlantic shortfin mako stock and comply with binding recommendations by the International Commission for the Conservation of Atlantic Tunas (ICCAT). 

```{r rec-landings, fig.cap = return_caption(chunk_name = "rec-landings", region = "NewEngland"),fig.width = 6.5,fig.asp = 0.4}
return_plot("rec-landings", region = "NewEngland")
```

```{r rec-hms, fig.cap = return_caption(chunk_name = "rec-hms", region = "NewEngland"), fig.width=6.5,fig.asp = 0.4, error=FALSE}
return_plot("rec-hms", region = "NewEngland")
```

### Implications

`r implications_text`

#### Stock Status

Single species [management objectives](https://noaa-edab.github.io/catalog/stock_status.html)  (1. maintaining biomass above minimum thresholds and 2. maintaining fishing mortality below overfishing limits) are not being met for some NEFMC managed species. Thirteen stocks are currently estimated to be below B~MSY~ (Fig. \ref{fig:stock-status}), while status relative to B~MSY~ could not be assessed for 13 additional stocks (Table \ref{tab:stock-status-table}). Although stock status and associated management constraints are likely contributing to decreased landings, current management constraints are in turn a response to biological conditions and not necessarily the primary cause of declining landings. To better address the role of management in future reports, we could examine how the total allowable catch (TAC) and the percentage of the TAC utilized for each species has changed through time.


```{r stock-status, fig.cap = return_caption(chunk_name = "stock-status", region = "NewEngland"), fig.width = 6, fig.asp = 0.8, cache = FALSE}
return_plot("stock-status", region = "NewEngland")
```

```{r stock-status-table, ft.arraystretch = 1}
stock_status_plot <- ecodata::plot_stock_status(report = "NewEngland")

stock_status_plot$unknown |>
flextable::flextable() |>
  flextable::set_header_labels(F.Fmsy = "F/Fmsy",
                               B.Bmsy = "B/Bmsy") |>
  flextable::colformat_num(na_str = "-") |>

  flextable::set_caption("Unknown or partially known stock status for NEFMC and jointly managed species.") |>
  flextable::footnote(i = 1:2, j = 1,value =flextable::as_paragraph('The most recent cod assessment made stock status recommendations for the four new stocks (Eastern Gulf of Maine, Western Gulf of Maine, Georges Bank, and Southern New England) but were not available yet for this report.'),
                      ref_symbols = c('1'))|>

  flextable::autofit() 
```

#### System Biomass

[Aggregate biomass](https://noaa-edab.github.io/catalog/aggregate_biomass.html) trends derived from scientific resource surveys have been stable to increasing in both regions (Fig. \ref{fig:nefsc-biomass-gb} & Fig. \ref{fig:nefsc-biomass-gom}).The benthivores group spiked during the last decade, due to a large haddock recruitment, but appears to be returning to average levels. Planktivore biomass on GB continues to rise with the highest fall biomass observed since 1968. There are mixed trends in piscivores on GB, and increasing trends for planktivores across both regions and seasons  and benthos on GB in both seasons. The New Hampshire/Maine state survey time series is too short to estimate trends, while the Massachusetts state survey shows the increasing trend in planktivores in the fall but a decrease in piscivores in the spring and benthos in both seasons (Fig. \ref{fig:mass-biomass}). While managed species comprise varying proportions of aggregate biomass, trends in landings are not mirroring shifts in the overall trophic structure of survey-sampled fish and invertebrates. Therefore, major shifts in feeding guilds or ecosystem trophic structure are unlikely to be driving the decline in landings.

Declining landings are likely driven by the relative abundance of specific targeted species rather than major shifts in ecosystem trophic structure or feeding guilds. Scientific surveys show that [Aggregate biomass](https://noaa-edab.github.io/catalog/aggregate_biomass.html) has been mostly stable or increasing in both regions (Fig. \ref{fig:nefsc-biomass-gb} & Fig. \ref{fig:nefsc-biomass-gom}). The benthivores(bottom-feeder) biomass recently peaked due to a large haddock recruitment, but appears to be returning to average levels. Planktivore (plankton-eater) biomass on GB continues to rise. On GB, trends in piscivores (fish-eaters) on GB are mixed, and benthos are increasing in both seasons. State-level data show the Massachusetts survey mirroring the increase in fall planktivores but noting a spring decrease in fish-eaters and a year-round decline in benthos; the New Hampshire/Maine survey remains too short to establish definitive trends (Fig. \ref{fig:mass-biomass}). 

```{r nefsc-biomass-gb, fig.cap = return_caption(chunk_name = "nefsc-biomass-gb", region = "NewEngland"), fig.width=6.5, fig.asp = 1,}
return_plot("nefsc-biomass-gb", region = "NewEngland")
```

```{r nefsc-biomass-gom, fig.cap = return_caption(chunk_name = "nefsc-biomass-gom", region = "NewEngland"),  fig.width=6.5, fig.asp = 1}
return_plot("nefsc-biomass-gom", region = "NewEngland")
```

#### Effect on Seafood Production

With the poor or unknown stock status of many managed species, the decline in commercial seafood landings in the Gulf of Maine most likely reflects lower catch quotas implemented to rebuild overfished stocks, as well as market dynamics.

The decline in recreational seafood harvest stems from multiple drivers. Some of the decline, such as for recreational shark landings, continues to be driven by tightening regulations. However, changes in demographics and preferences for recreational activities likely play a role in non-HMS (Highly Migratory Species) declines in recreational harvest, with current harvests well below the time series average. Recreational fishing may be shifting to catch-and-release strategies as opposed to catch for harvest.

```{r mass-biomass, fig.cap = return_caption(chunk_name = "mass-biomass", region = "NewEngland"), fig.width=6.5, fig.asp = 0.75}
return_plot("mass-biomass", region = "NewEngland")
```

Fisheries and communities rely on different combinations of stocks,and individual stocks will respond differently to these drivers. Some key drivers include :
  
  - Climatological conditions are trending into uncharted territory. Globally, 2025 had the warmest ocean temperatures on record^[https://noaa-edab.github.io/catalog/observation_synthesis.html] (see [2025 Highlights section](#highlights)). However, the 2025 Northwest Atlantic water temperatures were in line with the long-term average.
  - Stocks are shifting their distribution, moving towards the northeast and into deeper waters throughout the Northeast US Large Marine Ecosystem (Fig. \ref{fig:species-dist}, [Climate Risks section](#climate-risks)).
  - Some ecosystem composition and production changes have been observed (see [Stability section](#stability) and Risks to Setting Catch Limits section).
  - Some fishing communities are affected by Social and Community Risks (see [Social Vulnerability section](#social-vulnerability)).
  
<!-- \newpage -->

```{r, eval = params$id_child_docs, results = "asis"}
<<child-footer>>
```