```{r, eval = params$id_child_docs}
<<child-header>>
```

### Indicators: revenue (a proxy for profits)  
Total [commercial revenue](https://noaa-edab.github.io/catalog/comdat.html) and MAFMC managed species revenue within the Mid-Atlantic Bight have declined over the past 20-30 years. In 2023, total revenue was at an all-time low, and revenue from MAFMC managed species was near an all-time low (Fig. \ref{fig:comm-revenue}). 

```{r comm-revenue, fig.width = 6, fig.asp = 0.45, fig.cap = return_caption(chunk_name = "comm-revenue")}
```

Revenue earned by harvesting resources is a function of both the quantity landed of each species and the prices paid for landings. Beyond monitoring yearly changes in revenue, it is even more valuable to determine what drives these changes: harvest levels, the mix of species landed, price changes, or a combination of these. The [Bennet Indicator](https://noaa-edab.github.io/catalog/bennet.html) decomposes revenue change into two parts, one driven by changing quantities (volumes), and a second driven by changing prices. All changes are in relation to a base year (1982). The 1982 base year was selected because that is the first year the relevant data is available and it also allows for an extended period of time in which to evaluate market trends and dynamics.

In the Mid-Atlantic region revenues were above the 1982 baseline for all years in the series until 2022 and 2023 (Fig. \ref{fig:bennet}).  In 2023, lower revenue was driven primarily by both lower quantities of benthos landed, and lower prices of benthos, benthivores, and planktivores. The lower benthos prices are a departure from past years, which saw benthos prices contributing positively to revenue changes since the early 2000’s. (Fig.  \ref{fig:bennet-all}). 

```{r bennet, fig.cap = return_caption(chunk_name = "bennet"), fig.width=6, fig.asp=.35}
```

```{r bennet-all, fig.cap = return_caption(chunk_name = "bennet-all"),  fig.width = 6.5, fig.asp = 0.35}
bennet_all_plot
```

For ports combined across Mid-Atlantic states, [total climate vulnerability](https://noaa-edab.github.io/catalog/community_climate_vulnerability.html) of revenue ranged from high to very high from 2000-2021, with no long-term trend. This suggests that Mid-Atlantic port commercial fishing revenue has been highly reliant on climate-sensitive species for most of the period since 2000 (Fig. \ref{fig:climatevul-rev}). 

```{r climatevul-rev, fig.width=6.5, fig.asp=0.5, fig.cap=return_caption(chunk_name = "climatevul-rev")}
climatevul_rev_plot
```

### Implications 
Although the Mid-Atlantic region shows declining revenue since 2016, inflation-adjusted revenue from harvested species was still greater than 1982 levels until the past two years. In a similar manner to seafood landings, the results here are driven in large part by market dynamics affecting the landings of surfclams and ocean quahogs, as landings have been below quotas for these species, as well as lower quotas and prices for Atlantic scallops. The declining benthos category since 2012 may be partially caused by decreases in surfclam and ocean quahogs in the southern part of their range as harvest have shifted northward. Changes in other indicators, particularly those driving landings and those related to climate change, require monitoring as they may become important drivers of revenue in the future; for example:

-  Surfclams, ocean quahogs, and scallops are sensitive to warming ocean temperatures and ocean acidification, as reflected in the high climate vulnerability of total landings from from Mid-Atlantic ports. 
-  Multiple stressors including [warming](https://noaa-edab.github.io/catalog/bottom_temp_insitu.html) and [ocean acidification](https://noaa-edab.github.io/catalog/ocean_acidification) are interacting in Mid-Atlantic shellfish habitats.

```{r, eval = params$id_child_docs}
<<child-footer>>
```

