```{r, eval = FALSE, results = "asis"}
<<child-header>>
```

```{r, eval = params$draft}
<<ma-draft>>
```

### Indicators: revenue & net revenue  

Total [commercial revenue](https://noaa-edab.github.io/catalog/comdat.html) and MAFMC managed species revenue (2024 USD) within the Mid-Atlantic Bight have declined over the past 20 years. In 2024, total revenue and MAFMC managed species revenue were both near an all-time low (Fig. \ref{fig:comm-revenue}).

```{r comm-revenue, fig.width = 6, fig.asp = 0.45, fig.cap = return_caption(chunk_name = "comm-revenue", region = "MidAtlantic")}
return_plot("comm-revenue", region = "MidAtlantic")
```

Revenue earned by harvesting resources is a function of both the quantity landed of each species and the prices paid for landings. Therefore, total revenue patterns can be driven by harvest levels, the mix of species landed, price changes, or a combination of these. The [Bennet Indicator](https://noaa-edab.github.io/catalog/bennet.html) (BI) decomposes revenue change into two parts, one driven by changing quantities (volumes), and a second driven by changing prices. All changes are in relation to a base year (1982). The 1982 base year was selected because that is the first year the relevant data is available and it allows for an extended period of time to evaluate market trends and dynamics. The BI results demonstrate that relatively high revenues in 2014-2016 were somewhat equally due to higher landings and prices (Fig. \ref{fig:bennet}). In more recent years, both landings and prices have been closer to values from the reference year (1982). Low prices coupled with low volumes landed led to low revenue in 2024. Recent lower than baseline revenues are partially due to lower volumes and prices of benthos species (Fig. \ref{fig:bennet-all}).

```{r bennet, fig.cap = return_caption(chunk_name = "bennet", region = "MidAtlantic"), fig.width=6, fig.asp=.35}
return_plot("bennet", region = "MidAtlantic")
```

```{r bennet-all, fig.cap = return_caption(chunk_name = "bennet-all", region = "MidAtlantic"),  fig.width = 6.5, fig.asp = 0.35}
return_plot("bennet-all", region = "MidAtlantic")
```

```{r comdat-profit, fig.cap = return_caption(chunk_name = "comdat-profit", region = "MidAtlantic"),  fig.width = 6.5, fig.asp = 0.35}
return_plot("comdat-profit", region = "MidAtlantic")
```

This year, we present new indicators of [profitability](https://noaa-edab.github.io/catalog/comdat_profit.html): indices of cost, revenue, and profit based on trips catching federally-managed species. In this index, costs pertain to trip costs, excluding labor, estimated for all federal trips in the region. The profit indicator is net-revenue, determined as the difference between trip revenue and trip costs. Trips were spatially allocated to compile regional indices. Indices are presented as values relative to 2000, the first year in the dataset. In the Mid-Atlantic, costs have fluctuated, but overall remain near the time series mean, despite some high costs in 2022, 2014 and 2008. The revenue index, however, has declined steadily since 2019 and is driving an overall decline in profits  (Fig. \ref{fig:comdat-profit}). 

For Mid-Atlantic ports, [total vulnerability](https://noaa-edab.github.io/catalog/community_climate_vulnerability.html) of revenue is high for the entire time series [(2000-2024)](#social-and-community-risks), with no long-term trend. This suggests that Mid-Atlantic port commercial fishing revenue is highly reliant on climate-sensitive species for the entire time period assessed (Fig. \ref{fig:climatevul-rev}). 

```{r climatevul-rev, fig.width=6.5, fig.asp=0.5, fig.cap=return_caption(chunk_name = "climatevul-rev", region = "MidAtlantic")}
return_plot("climatevul-rev", region = "MidAtlantic")
```

### Implications 
Although the Mid-Atlantic region shows declining revenue since 2016 (Fig. \ref{fig:comm-revenue}), inflation-adjusted revenue from harvested species was still greater than 1982 levels until the past two years (Fig. \ref{fig:bennet}). However, revenue from MAFMC-managed species has been below 2000 levels in several of the past 24 years (Fig. \ref{fig:comdat-profit}). The Bennet Index demonstrates that this decline is driven by lower landing volumes, which were not offset by price increases. Declines in landings of surfclams and ocean quahogs since 2012 may reflect changes in surfclam and quahog aggregation or distribution patterns and/or changes to [landings per unit effort](https://noaa-edab.github.io/catalog/effective_sweptarea.html). Changes in other indicators, particularly those driving landings and those related to climate change, require monitoring as they may become important drivers of revenue in the future.  Multiple stressors including [warming](https://noaa-edab.github.io/catalog/bottom_temp_insitu.html) and [ocean acidification](https://noaa-edab.github.io/catalog/ocean_acidification) are interacting in Mid-Atlantic shellfish habitats, particularly for surfclams, ocean quahogs, and scallops. This is reflected by the high environmental risk for landings from Mid-Atlantic ports.

```{r, eval = FALSE, results = "asis"}
<<child-footer>>
```

