```{r, eval = FALSE, results = "asis"}
<<child-header>>
```

```{r, eval = params$draft}
<<ma-draft>>
```

### Indicators: revenue & net revenue  

Total [commercial revenue](https://noaa-edab.github.io/catalog/comdat.html) and MAFMC managed species revenue (2024 USD) within the Mid-Atlantic Bight have declined over the past 20 years. Despite these declines, inflation-adjusted revenue from harvested species was still greater than 1982 levels until the past two years (Fig. \ref{fig:bennet}). However, revenue from MAFMC-managed species has been below 2000 levels in several of the past 24 years (Fig. \ref{fig:comdat-profit}). In 2024, total revenue and MAFMC managed species revenue were both near an all-time low (Fig. \ref{fig:comm-revenue}). Declines in landings of surfclams and ocean quahogs since 2012 may reflect changes in surfclam and quahog aggregation or distribution patterns and/or changes to [landings per unit effort](https://noaa-edab.github.io/catalog/effective_sweptarea.html).

```{r comm-revenue, fig.width = 6, fig.asp = 0.45, fig.cap = return_caption(chunk_name = "comm-revenue", region = "MidAtlantic")}
return_plot("comm-revenue", region = "MidAtlantic")
```

```{r comdat-profit, fig.cap = return_caption(chunk_name = "comdat-profit", region = "MidAtlantic"),  fig.width = 6.5, fig.asp = 0.35}
return_plot("comdat-profit", region = "MidAtlantic")
```

This year, we present new indicators of [profitability](https://noaa-edab.github.io/catalog/comdat_profit.html): indices of cost, revenue, and profit based on trips catching federally-managed species. In the Mid-Atlantic, costs have fluctuated, but overall remain near the time series mean, despite some high costs in 2022, 2014 and 2008. The revenue index, however, has declined steadily since 2019 and is driving an overall decline in profits (Fig. \ref{fig:comdat-profit}). 

### Drivers: Bennet Index

Revenue earned by harvesting resources is a function of both the quantity landed of each species and the prices paid for landings. Therefore, total revenue patterns can be driven by harvest levels, the mix of species landed, price changes, or a combination of these. The [Bennet Indicator](https://noaa-edab.github.io/catalog/bennet.html) (BI) decomposes revenue change into two parts, one driven by changing quantities (volumes), and a second driven by changing prices. Relatively high revenues in 2014-2016 were somewhat equally due to higher landings and prices (Fig. \ref{fig:bennet}). In more recent years, both landings and prices have been closer to values from the reference year (1982). Low prices coupled with low volumes landed led to low revenue in 2024. Recent lower than baseline revenues are partially due to lower volumes and prices of benthos species (Fig. \ref{fig:bennet-all}). OVerall, lower landing volumes have not been offset by price increases.  

```{r bennet, fig.cap = return_caption(chunk_name = "bennet", region = "MidAtlantic"), fig.width=6, fig.asp=.35}
return_plot("bennet", region = "MidAtlantic")
```

```{r bennet-all, fig.cap = return_caption(chunk_name = "bennet-all", region = "MidAtlantic"),  fig.width = 6.5, fig.asp = 0.35}
return_plot("bennet-all", region = "MidAtlantic")
```

### Implications 

For Mid-Atlantic ports, [total vulnerability](https://noaa-edab.github.io/catalog/community_risks.html) of revenue is high for the entire time series [(2000-2024)](#social-and-community-risks), with no long-term trend (Fig. \ref{fig:climatevul-rev}), reflecting reliance on shellfish, which are sentitive to multiple environmental stressors including [warming](https://noaa-edab.github.io/catalog/bottom_temp_insitu.html) and [ocean acidification](https://noaa-edab.github.io/catalog/ocean_acidification).

```{r climatevul-rev, fig.width=6.5, fig.asp=0.5, fig.cap=return_caption(chunk_name = "climatevul-rev", region = "MidAtlantic")}
return_plot("climatevul-rev", region = "MidAtlantic")
```

```{r, eval = FALSE, results = "asis"}
<<child-footer>>
```

