```{r, eval = params$id_child_docs, results = "asis"}
<<child-header>>
```

```{r, eval = params$draft}
<<ma-draft>>
```

### Indicators: revenue (a proxy for profits)  

Total [commercial revenue](https://noaa-edab.github.io/catalog/comdat.html) and MAFMC managed species revenue (2024 USD) within the Mid-Atlantic Bight have declined over the past 20 years. In 2024, total revenue and MAFMC managed species revenue were both near an all-time low (Fig. \ref{fig:comm-revenue}).

```{r comm-revenue, fig.width = 6, fig.asp = 0.45, fig.cap = return_caption(chunk_name = "comm-revenue", region = "MidAtlantic")}
return_plot("comm-revenue", region = "MidAtlantic")
```

Revenue earned by harvesting resources is a function of both the quantity landed of each species and the prices paid for landings. Therefore, total revenue patterns can be driven by harvest levels, the mix of species landed, price changes, or a combination of these. The [Bennet Indicator](https://noaa-edab.github.io/catalog/bennet.html) (BI) decomposes revenue change into two parts, one driven by changing quantities (volumes), and a second driven by changing prices. All changes are in relation to a base year (1982). The 1982 base year was selected because that is the first year the relevant data is available and it allows for an extended period of time to evaluate market trends and dynamics. The BI results demonstrate that relatively high revenues in 2014-2016 were equally due to higher landings and prices (Fig. \ref{fig:bennet}). In more recent years, both landings and prices have been closer to values from the reference year (1982). A low year for prices in 2024, coupled with low volumes landed, led to low revenue. Recent lower than average revenues are partially due to declining prices of benthivores. Benthos prices increased from 2023, but overall benthos revenue remained low due to low volumes landed (Fig.  \ref{fig:bennet-all}).

```{r bennet, fig.cap = return_caption(chunk_name = "bennet", region = "MidAtlantic"), fig.width=6, fig.asp=.35}
return_plot("bennet", region = "MidAtlantic")
```

```{r bennet-all, fig.cap = return_caption(chunk_name = "bennet-all", region = "MidAtlantic"),  fig.width = 6.5, fig.asp = 0.35}
return_plot("bennet-all", region = "MidAtlantic")
```

This year, we present new indicators of profitability: indices of cost, revenue, and profit based on trips catching federally-managed species. In this index, costs pertain to trip costs, excluding labor, estimated for all federal trips in the region using methods described in Werner et al. (2020). The profit indicator is net-revenue, determined as the difference between trip revenue and trip costs. Trips were spatially allocated to compile regional indices. Indices are presented as values relative to 2000, the first year in the dataset. In the Mid-Atlantic, costs have fluctuated, but overall remain near the time series mean, despite some high costs in 2022, 2014 and 2008. Revenue, however, has declined steadily since 2019 and is driving an overall decline in profits (Fig. \ref{fig:comdat-profit}). 

```{r comdat-profit, fig.cap = return_caption(chunk_name = "comdat-profit", region = "MidAtlantic"),  fig.width = 6.5, fig.asp = 0.35}
return_plot("comdat-profit", region = "MidAtlantic")
```

For ports combined across Mid-Atlantic states, [total climate vulnerability](https://noaa-edab.github.io/catalog/community_climate_vulnerability.html) of revenue ranged from high to very high from 2000-2021, with no long-term trend. This suggests that Mid-Atlantic port commercial fishing revenue has been highly reliant on climate-sensitive species for most of the period since 2000 (Fig. \ref{fig:climatevul-rev}). 

```{r climatevul-rev, fig.width=6.5, fig.asp=0.5, fig.cap=return_caption(chunk_name = "climatevul-rev", region = "MidAtlantic")}
return_plot("climatevul-rev", region = "MidAtlantic")
```

### Implications 
Although the Mid-Atlantic region shows declining revenue since 2016, inflation-adjusted revenue from harvested species was still greater than 1982 levels until the past two years (Fig. \ref{fig:bennet}). However, revenue from MAFMC-managed species has been below 2000 levels in several of the past 24 years (Fig.  \ref{fig:comdat-profit}). The BI demonstrates that this decline is driven by lower volumes and no inverse price effects to offset the decreases in volume. Declines in landings of surfclams and ocean quahogs since 2012 are a result of decreased landings per unit effort over the same period, which may reflect changes in surfclam and quahog aggregation or distribution patterns. Changes in other indicators, particularly those driving landings and those related to climate change, require monitoring as they may become important drivers of revenue in the future; for example:

-  Surfclams, ocean quahogs, and scallops are sensitive to warming ocean temperatures and ocean acidification, as reflected in the high climate vulnerability of total landings from from Mid-Atlantic ports. 
-  Multiple stressors including [warming](https://noaa-edab.github.io/catalog/bottom_temp_insitu.html) and [ocean acidification](https://noaa-edab.github.io/catalog/ocean_acidification) are interacting in Mid-Atlantic shellfish habitats.

```{r, eval = params$id_child_docs, results = "asis"}
<<child-footer>>
```

