```{r, eval = FALSE, results = "asis"}
<<child-header>>
```

```{r, eval = params$draft}
<<ma-draft>>
```

### Indicators: Angler trips, fleet diversity  
[Recreational effort](https://noaa-edab.github.io/catalog/recdat.html) (angler trips) in the MAB increased from 1982 to 2010, but has since declined to near the long-term average (Fig. \ref{fig:rec-op}). in the MAB. MAB. However, there is a long-term declining trend in recreational fleet diversity (i.e., effort by shoreside, private boat, and for-hire anglers) (Fig. \ref{fig:rec-div}). Billfish landings were notably high in 2025 (See [2025 Highlights](https://noaa-edab.github.io/catalog/observation_synthesis_2025.html) Section), but long-term time series are in development.

```{r rec-op, fig.cap = return_caption(chunk_name = "rec-op", region = "MidAtlantic")}
return_plot("rec-op", region = "MidAtlantic")
```

```{r rec-div, fig.cap = return_caption(chunk_name = "rec-div", region = "MidAtlantic")}
return_plot("rec-div", region = "MidAtlantic")
```

### Implications
While the overall number of recreational trips in the MAB is above the long-term average, the continuing decline in recreational fleet effort diversity suggests, at least in part, changes in angler behavior. Future study is required to determine whether and to what extent the range and availability of recreational fishing options may drive these changes as well.

A contraction of party/charter trips (dropping from 2.2% of trips in 2021 to 1.3% in 2023) is the primary driver of the downward effort diversity trend, alongside a shift toward shorebased angling, which now makes up 60% of trips. Private boat effort has remained consistent to 2022 values.

Managers should consider the differing species and fish sizes for shore-based and vessel-based anglers. Some species use inshore regions as nursery grounds while other species only come inshore as adults. Many states have developed shore-based regulations where the minimum size is lower than in other areas and sectors to maintain opportunities in the shore angling sector. The MAFMC is currently considering recreational sector separation which might establish different options for managing the for-hire sector from other modes.


```{r, eval = FALSE, results = "asis"}
<<child-footer>>
```