```{r, eval = params$id_child_docs, results = "asis"}
<<child-header>>
```

### Indicators: Angler trips, fleet diversity  

[Recreational effort](https://noaa-edab.github.io/catalog/recdat.html) (angler trips) increased during 1982-2010, but has since declined to the long-term average (Fig. \ref{fig:rec-op}). Recreational fleets are defined as private vessels, shore-based fishing, or party-charter vessels. Recreational fleet diversity, or the relative importance of each fleet type, has remained relatively stable over the latter half of the time series (Fig.  \ref{fig:rec-div}).
```{r rec-op, fig.cap = return_caption(chunk_name = "rec-op"),fig.width = 6.5,fig.asp = 0.3}
rec_op_plot
```

```{r rec-div, fig.cap = return_caption(chunk_name = "rec-div"),fig.width = 6.5,fig.asp = 0.3}
rec_div_plot +
  ggplot2::ylab('Effective Shannon Index')
```

### Implications

The absence of a long term trend in recreational angler trips and fleet effort diversity suggests relative stability in the overall number of recreational opportunities in the region.

```{r, eval = params$id_child_docs, results = "asis"}
<<child-footer>>
```