```{r, eval =FALSE, results = "asis"}
<<child-header>>
```

```{r, eval = params$draft}
<<ne-draft>>
```

### Indicators: Angler trips, fleet diversity  

[Recreational effort](https://noaa-edab.github.io/catalog/recdat.html) (angler trips) increased from 1982 to 2010, but has since declined to just below the long-term average (Fig. \ref{fig:rec-op-ne}). Recreational fleets are defined as private vessels, shore-based fishing, or party-charter vessels. Recreational fleet diversity, or the relative importance of each fleet type, has remained relatively stable over the latter half of the time series (Fig.  \ref{fig:rec-div-ne}). Billfish landings were notably high in 2025 (See [2025 Highlights](https://noaa-edab.github.io/catalog/observation_synthesis_2025.html) Section), but long-term time series are in development.

```{r rec-op-ne, fig.cap = return_caption(chunk_name = "rec-op", region = "NewEngland"),fig.width = 6.5,fig.asp = 0.3}
return_plot("rec-op", region = "NewEngland")
```

```{r rec-div-ne, fig.cap = return_caption(chunk_name = "rec-div", region = "NewEngland"),fig.width = 6.5,fig.asp = 0.3}
return_plot("rec-div", region = "NewEngland")
```

### Implications

The absence of a long term trend in recreational angler trips and fleet effort diversity suggests relative stability in the overall number of recreational opportunities in the region.

```{r, eval = FALSE, results = "asis"}
<<child-footer>>
```