```{r, eval = params$id_child_docs, results = "asis"}
<<child-header>>
```

```{r, eval = params$draft}
<<ma-draft>>
```

### Indicators: Angler trips, fleet diversity  
[Recreational effort](https://noaa-edab.github.io/catalog/recdat.html) (angler trips) in 2023 continues to be above the long-term average (Fig. \ref{fig:rec-op}). in the MAB. MAB. However, there is a long-term declining trend in recreational fleet diversity (i.e., effort by shoreside, private boat, and for-hire anglers) (Fig. \ref{fig:rec-div}). Billfish landings were notably high in 2025 (See 2025 Highlights Section), but long-term time series are in development.

```{r rec-op, fig.cap = return_caption(chunk_name = "rec-op", region = "MidAtlantic")}
return_plot("rec-op", region = "MidAtlantic")
```

```{r rec-div, fig.cap = return_caption(chunk_name = "rec-div", region = "MidAtlantic")}
return_plot("rec-div", region = "MidAtlantic")
```

### Implications
While the overall number of recreational trips in the MAB is above the long-term average, the continuing decline in recreational fleet effort diversity suggests, at least in part, changes in angler behavior. Future study is required to determine whether and to what extent the range and availability of recreational fishing options may drive these changes as well.

The downward effort diversity trend is driven by party/charter contraction (down from 2.2% in 2021 to 1.3% of trips in 2023), and a shift toward shorebased angling, which currently makes up 60% of all angler trips. Private boat effort has remained relatively stable compared to 2022 values.

Shore anglers will have access to different species than vessel-based anglers, and when the same species is accessible both from shore and from a vessel, fish size differs by location, with some species using inshore regions as nursery grounds while other species only come inshore as adults. Many states have developed shore-based regulations where the minimum size is lower than in other areas and sectors to maintain opportunities in the shore angling sector. MAFMC is currently considering recreational sector separation which might establish different options for managing the for-hire sector from other modes.

```{r, eval = params$id_child_docs, results = "asis"}
<<child-footer>>
```