```{r, eval = FALSE, results = "asis"}
<<child-header>>
```

```{r, eval = params$draft}
<<ma-draft>>
```

#### Fishery Stability 
Indicators suggest that Mid-Atlantic fisheries have broadly shifted from the historic baseline. Commercial fishery fleet count has declined while fleet revenue diversity has been stable over time in the MAB, but current fleet revenue diversity is above the long-term average (Fig. \ref{fig:comm-div-fleet}). This indicates that the commercial fleet composition has changed, but the portfolio of species targeted is similar over time (Fig. \ref{fig:commercial-div-species-div}). [Revenue per unit effort](https://noaa-edab.github.io/catalog/effective_sweptarea.html) remains steady or increasing over time for most gear types, indicating financial viability of current fishing operations. Target species such as Atlantic mackerel and quahog have had reduced catch limits in recent years, resulting in reduced landings in these fisheries, and a decline in scallop catch within the MAB has severely reduced the total revenue generated in the region. Because non-MAFMC managed landings and revenue have declined, a larger share of the regional landings and revenue come from Council-managed fisheries.

The [Crew Survey](https://noaa-edab.github.io/catalog/crew_survey.html) shows that specific aspects pertaining to sustainability and resilience of the fishing lifestyle are declining: predictability of earnings, the amount of time away from home, the physical fatigue of the job, and the personal health impacts have all been cited as dissatisfaction rates increase. Overall job satisfaction remains relatively stable over time, but unveils vulnerability as additional survey results show an aging population, particularly an increase in the 55+ crew cohort, and fewer individuals entering the fishery. This suggests a reduced capacity for Mid-Atlantic commercial fisheries to adapt to future uncertainties and change. New [Communities at Sea](https://noaa-edab.github.io/catalog/communities_at_sea.html) indicators that assess fishing communities’ ability to adapt to change are in development and will provide additional fishing industry indicators in future reports.

Despite reduced [recreational landings](https://noaa-edab.github.io/catalog/recdat.html) (Fig. \ref{fig:rec-landings}), the number of recreational trips is near average (Fig. \ref{fig:rec-op}), suggesting a shift to catch-and-release fishing. Billfish (i.e., white marlin) catch-and-release was especially high, possibly due to shifting effort due to the closure of the recreational bluefin tuna fishery in August 2025. Shark and large sport fish regulations and environmental conditions, among other circumstances, may also contribute to reduced recreational landings. As noted above, [recreational fleet effort diversity](https://noaa-edab.github.io/catalog/recdat.html) is declining (Fig. \ref{fig:rec-div}), suggesting a shift in modes of recreational fishery participation. The Mid-Atlantic has experienced a contraction of the party and charter sectors, with more recreational angling occurring from shore. Recreational species catch diversity has no long-term trend and has been at or above the long-term average since 2016 (Fig. \ref{fig:recdat-div-catch}), indicating that anglers continue to catch a mix of species. 

```{r comm-div-fleet, fig.width = 6.5, fig.asp = 0.3, fig.cap=return_caption(chunk_name = "comm-div-fleet", region = "MidAtlantic")}
return_plot("comm-div-fleet", region = "MidAtlantic")
```


```{r commercial-div-species-div,  fig.cap = return_caption(chunk_name = "commercial-div-species-div", region = "MidAtlantic")}
return_plot("commercial-div-species-div", region = "MidAtlantic")
```


```{r recdat-div-catch, fig.cap =return_caption(chunk_name = "recdat-div-catch", region = "MidAtlantic")}
return_plot("recdat-div-catch", region = "MidAtlantic")
```

#### Ecological Stability
Long-term changes in biological processes suggest the Mid-Atlantic ecosystem is experiencing a systemic shift. Total annual [primary production](https://noaa-edab.github.io/catalog/chl_pp.html), a measure of the total amount of carbon (i.e., energy) produced by phytoplankton per year, has no clear trend (Fig. \ref{fig:totpp}), suggesting stability in energy at the base of the food web. However, we are monitoring for shifts in the phytoplankton community, which can affect the amount of primary production available to higher trophic levels. [Zooplankton diversity](https://noaa-edab.github.io/catalog/zoo_diversity.html) is increasing in the MAB, and measures of zooplankton community composition also indicate a long-term shift in [zooplankton communities](https://noaa-edab.github.io/catalog/zoo_community.html) (Fig. \ref{fig:zoo-community}). Together, these indicators show a gradual but systemic change in lower trophic levels towards a community with a higher proportion of euphausiids and less dominated by copepods, which would not be expected in a stable ecosystem.

There are long-term increases in the biomass of the euphausid, benthivore, and benthos [guilds](https://noaa-edab.github.io/catalog/aggregate_biomass.html) (Fig. \ref{fig:nefsc-biomass-mab}). These lower trophic groups have similar roles within the ecosystem and these changes indicate a shift towards an ecosystem with a higher representation of those functional groups. [Adult fish diversity](https://noaa-edab.github.io/catalog/exp_n.html), the expected number of species in a standard number of individuals sampled from the NEFSC bottom trawl survey, had no trend when the Albatross vessel was used for sampling, but shows a short-term declining trend since switching to the Bigelow vessel in 2008. However, the expected number of species observed is currently within one standard deviation from most historic estimates (Fig. \ref{fig:exp-n}). This suggests that biomass increases in some guilds is due to an overall producivity increase rather than an influx of new species.


```{r totpp, fig.cap=return_caption(chunk_name = "totpp", region = "MidAtlantic")}
return_plot("totpp", region = "MidAtlantic")
```


<!-- # ```{r zoo-diversity, fig.cap = return_caption(chunk_name = "zoo-diversity", region = "MidAtlantic"), fig.asp=.47} -->
<!-- # return_plot("zoo-diversity", region = "MidAtlantic") -->
<!-- # ``` -->
<!-- #  -->

```{r exp-n, fig.cap = return_caption(chunk_name = "exp-n", region = "MidAtlantic"), fig.width=5, fig.asp=.4}
return_plot("exp-n", region = "MidAtlantic")
```

```{r zoo-community, fig.cap = return_caption(chunk_name = "zoo-community", region = "MidAtlantic"), fig.width=5, fig.asp=.4}
return_plot("zoo-community", region = "MidAtlantic")
```

<!-- \newpage -->

[Functional traits](https://noaa-edab.github.io/catalog/finfish_traits.html), such as length at maturity, maximum body size, and fecundity, serve to synthesize change in complex, diverse communities by looking beyond species-specific trends. Furthermore, shifts in functional trait distributions for the fish community can indicate changes in  ecosystem-scale resilience. There is evidence of long-term change in trait distributions in the MAB, particularly in the fall season (Fig. \ref{fig:traits-k}, \ref{fig:traits-tl}). The fall finfish community in the MAB is showing long-term shifts towards faster life history strategies with lower trophic levels, smaller offspring, younger age and shorter length at maturity, and faster growth rates.  This indicates shifts in a system increasingly composed of smaller, fast-growing species. The long-term trends in the spring season are, however, more equivocal, with some evidence in shifts towards slower life history strategies, including larger length-at-maturity and offspring size. The lack of trend in finfish diversity suggests that these changes in fish communities are not due to a change in the total number of species.

```{r traits-k, fig.width = 5, fig.asp=0.5, fig.cap= return_caption(chunk_name = "traits-k", region = "MidAtlantic")}
return_plot("traits-k", region = "MidAtlantic")
```

```{r traits-tl, fig.width = 5, fig.asp=0.5, fig.cap= return_caption(chunk_name = "traits-tl", region = "MidAtlantic")}
return_plot("traits-tl", region = "MidAtlantic")
```

### Implications
Fleet diversity indices are used by the MAFMC in their EAFM risk assessment to evaluate stability objectives, as well as risks to fishery resilience and maintaining equity in access to fishery resources. Instability in the commercial fleet count metric suggests lower capacity to respond to the current range of fishing opportunities. Commercial species permit revenue diversity is relatively stable (Fig. \ref{fig:commercial-div-species-div}) but comparisons are limited by missing historical (pre-2003) clam fishery data.

Declining recreational fleet effort diversity indicates that the party/charter boat sector continues to contract, with shoreside angling becoming a greater percentage of recreational angler trips. Stability in recreational species catch diversity has been maintained by a different set of species over time. Since the 1990s, there has been an increase in Atlantic States Marine Fisheries Commission (ASMFC) and South Atlantic Fishery Management Council (SAFMC) managed species in recreational catch, concomitant with a decline in MAFMC and New England Fishery Management Council (NEFMC) managed species. These changes in effort and species trends may necessitate new or changing management considerations to ensure effective tools and opportunities are in place to support recreational fisheries.

Production at the base of the food web is variable but stable over time. Mid-Atlantic species composition is changing, shifting towards a higher proportion of benthic and demersal fish. Stable adult fish diversity indicates the same overall number and evenness over time, but doesn’t rule out species substitutions (e.g., warm-water species replacing cold-water species). There is evidence for long-term change in finfish trait distributions in the Mid-Atlantic. Future research should consider how changes in single species abundances may influence ecosystem stability.

In the MAB, both the fisheries and ecosystem are exhibiting long-term systemic shifts away from historical norms. While these changes don’t appear abrupt like one would expect during a regime shift, they do indicate a potential change in baseline conditions.


```{r, eval = FALSE, results = "asis"}
<<child-footer>>
```