```{r, eval = params$id_child_docs, results = "asis"}
<<child-header>>
```

### Indicators: Fishing Engagement and Community Social Vulnerability

The [engagement indices](https://noaa-edab.github.io/catalog/engagement.html) demonstrate the importance of commercial and recreational fishing to a given community relative to other coastal communities in a region. Social vulnerability indicators measure social factors that shape a community’s ability to adapt to change.  

For this report, we focus on top communities with the highest engagement scores, the top communities with the highest population relative engagement scores, and on three socio-demographic indicators within the CSVI toolset (poverty, personal disruption, population composition). 

In 2022, New Bedford, MA stands out as having a particularly high engagement in commercial fishing, while Frenchboro, ME is much more engaged in commercial fishing relative to its population size (Fig. \ref{fig:commercial-engagement}). Of particular concern among top communities are New Bedford and Boston, MA as they both have medium or higher scores for all three socio-demographic indicators, while Port Clyde-Tenants Harbor, ME ranked medium for two of the three indicator (Table \ref{tab:comm-vuln}).

```{r commercial-engagement, fig.cap= return_caption(chunk_name = "commercial-engagement"), fig.width = 7, fig.asp = 0.5, results='hide'}
commercial_engagement_plot
```

```{r comm-vuln}
comm_vul_table |>
  flextable::fit_to_width(7.5)
```

Narragansett/Point Judith, RI; Newington, NH; and Gloucester, MA ranked as top communities for both commercial and recreational indices (Fig. \ref{fig:recreational-engagement}), suggesting that they may be impacted simultaneously (to a greater degree than others) by commercial and recreational regulatory changes. Of the top-ranked recreational communities, only Provincetown, MA and Falmouth, MA had medium or higher ranks for more than one socio-demographic indicator (Table. \ref{tab:rec-vuln}) examined here (poverty, personal disruption, population composition). This suggests that future changes to recreational fishing conditions may disproportionately impact Provincetown and Falmouth.

\newpage
```{r recreational-engagement, fig.cap= return_caption(chunk_name = "recreational-engagement"), fig.width = 6.5, fig.asp = 0.7, results='hide'}
recreational_engagement_plot
```
\newpage

```{r rec-vuln}
rec_vul_table|>
  flextable::fit_to_width(7.5)
```

### Indicators: Community Climate Vulnerability in the New England

The [Community Climate Change Risk Indicators](https://noaa-edab.github.io/catalog/community_climate_vulnerability.html) are calculated by multiplying the percent contribution of species to the total value landed in a community by their respective Total Vulnerability scores (based on NOAA’s Climate Vulnerability Assessment) for different sensitivity and exposure factors and then summing the resulting values by year. As a community (or region) shifts towards climate vulnerable species, its risk score increases. While there is not a long-term trend in total climate vulnerability across New England communities as a whole, the proportion of communities with moderate vulnerability is decreasing and shifting more towards high or very high vulnerability scores (Fig. \ref{fig:commvulex}). This suggests that some communities are shifting towards being more dependent on climate-vulnerable species, particularly shellfish.

```{r commvulex, fig.width = 6.5, fig.asp=.5, fig.cap=return_caption(chunk_name = "commvulex")}
commvulex_plot
```

### Implications
These indicators provide a snapshot of the presence of socio-demographic concerns in the most highly engaged commercial and recreational fishing communities in New England. These communities may be especially vulnerable to changes in fishing patterns due to regulations and/or ecosystem changes. Several of  these top fishing communities, both commercial and recreational fishing communities, demonstrated medium to high socio-demographic vulnerability, indicating that they may be at a disadvantage responding to change.

```{r, eval = params$id_child_docs, results = "asis"}
<<child-footer>>
```