```{r, eval = params$id_child_docs, results = "asis"}
<<child-header>>
```

```{r, eval = params$draft}
<<ma-draft>>
```

### Indicators: Port Commercial Fishing Activity and Community Social Vulnerability

Six of the top ten 2024 communities are below in the [Port Commercial Fishing Activity Indicator (PCFA)](https://noaa-edab.github.io/catalog/engagement.html?q=engage#introduction-to-indicator-72) compared to their average scores from 2007-2011: Point Pleasant Beach, NJ; Ocean City, MD; Bronx, NY; Barnegat Light, NJ; Newport News, VA; Cape May, NJ. Of particular concern, Atlantic City, NJ and Newport News, VA both rank medium or higher for all three socio-demographic CSVIs, suggesting that fishing industry participants associated with this municipality may be more vulnerable to change. The other four top communities show higher port activity since 2007-2011; most notably Hampton Bays/Shinnecock, NY with an increase of 84%. Currently North Carolina communities are not presented due to data limitations.

```{r commercial-engagement, fig.cap= return_caption(chunk_name = "commercial-engagement", region = "MidAtlantic"), fig.width = 6.5, fig.asp = 0.85}
return_plot("commercial-engagement", region = "MidAtlantic")
```


```{r comvultab}
comm_vul_table
```

Of those included in the top-ranked recreational communities, both Morehead City, NC and Virginia Beach, VA had medium or higher ranks for at least one socio-demographic indicator (Table \ref{tab:recvultab}). This suggests that future changes to recreational fishing conditions may disproportionately impact these places.

```{r recreational-engagement, fig.cap= return_caption(chunk_name = "recreational-engagement", region = "MidAtlantic"), fig.width = 6.5, fig.asp = 0.85}
return_plot("recreational-engagement", region = "MidAtlantic")
```


```{r recvultab}
rec_vul_table
```


### Indicators: Community Environmental Variability Risk  in the Mid-Atlantic

[Community Environmental Variability Risk Indicators](https://noaa-edab.github.io/catalog/community_climate_vulnerability.html) (CEVRI) measure risk by linking commercial landings and revenue to specific climate sensitivity factors, including temperature, ocean acidification, and stock status using the Climate Vulnerability Assessment (CVA) scores. These indicators calculate total sensitivity and vulnerability scores based on a community's dependence on species vulnerable to climate change. Risk scores range from low (1) to high (4), increasing as a community relies more heavily on species at higher risk from environmental shifts.

While long-term risk trends across the Mid-Atlantic remain stable, most individual fishing communities currently rank as high or very high risk. This high ranking demonstrates that a majority of regional communities depend on species that are highly vulnerable to changing ocean conditions for their commercial revenue. Strategies for management should account for this widespread reliance on climate-sensitive stocks.


```{r commvulex, fig.width = 6.5, fig.asp=.5, fig.cap=return_caption(chunk_name = "commvulex", region = "MidAtlantic")}
return_plot("commvulex", region = "MidAtlantic")
```


### Implications

A range of socioeconomic and environmental variability risk concerns are found throughout Mid-Atlantic fishing communities, and the CSVI and CEVRI indicate socio-demographic concerns in the most highly active commercial Mid-Atlantic fishing ports. Fishing industry participants that utilize more vulnerable ports may be at increased relative risk to changes in fishing patterns due to regulations and/or ecosystem changes.

A majority of Mid-Atlantic communities have high to very high total environmental variability risk based on revenue. Coastal fishing communities are greatly affected by environmental change, both because of their physical location and because of their frequent social, cultural, and economic dependence on fishing. These impacts are expected to become more pressing as changes become more extensive. Changes in [ocean temperature](https://noaa-edab.github.io/catalog/bottom_temp_insitu.html) and [acidification](https://noaa-edab.github.io/catalog/ocean_acidification.html) affecting marine life have the potential to directly impact fisheries and fishery dependent livelihoods.

```{r, eval = params$id_child_docs, results = "asis"}
<<child-footer>>
```