```{r, eval = params$id_child_docs, results = "asis"}
<<child-header>>
```

```{r, eval = params$draft}
<<ne-draft>>
```

### Indicators: Port Commercial Fishing Activity and Community Social Vulnerability

The Port Commercial Fishing Activity Indicator (PCFA) highlights significant shifts in industry engagement across major regional ports. New Bedford and Boston, MA, have experienced fishing activity declines of over 20% compared to their 2007–2011 averages. Portland, ME, and Gloucester, MA, show even sharper downturns, with activity dropping 39% and 45%, respectively. Because New Bedford and Boston also rank medium-high for socio-demographic vulnerability, industry participants in these municipalities face a higher risk from these changing conditions.Conversely, several communities show substantial growth in fishing activity. Chatham, MA, along with Stonington and Harpswell, ME, are seeing positive trends. Friendship, ME, stands out with a 95% increase in its PCFA score over the same period. 

```{r commercial-engagement-ne, fig.cap= return_caption(chunk_name = "commercial-engagement", region = "NewEngland"), fig.width = 7, fig.asp = 0.5}
return_plot("commercial-engagement", region = "NewEngland")
```

```{r comm-vuln}
comm_vul_table |>
  flextable::fit_to_width(7.5)
```

Of the top-ranked recreational communities, only Provincetown, MA and Falmouth, MA had medium or higher ranks for more than one socio-demographic indicator (Table \ref{tab:recvultab}) examined here (poverty, personal disruption, population composition). This suggests that future changes to recreational fishing conditions may disproportionately impact Provincetown and Falmouth.

\newpage
```{r recreational-engagement-ne, fig.cap= return_caption(chunk_name = "recreational-engagement", region = "NewEngland"), fig.width = 6.5, fig.asp = 0.7}
return_plot("recreational-engagement", region = "NewEngland")
```
<!-- \newpage -->

```{r rec-vuln}
rec_vul_table|>
  flextable::fit_to_width(7.5)
```

### Indicators: Community Environmental Variability Risk in New England

[Community Environmental Variability Risk Indicators](https://noaa-edab.github.io/catalog/community_climate_vulnerability.html) (CEVRI) measure risk by linking commercial landings and revenue to specific climate sensitivity factors, including temperature, ocean acidification, and stock status using the Climate Vulnerability Assessment (CVA) scores (based on Hare et al. 2016). These indicators calculate total sensitivity and vulnerability scores based on a community's dependence on species vulnerable to climate change. Risk scores range from low (1) to high (4), increasing as a community relies more heavily on species at higher risk from environmental shifts.While there is no long-term trend in risk across New England communities,  the proportion of communities with moderate risk is decreasing and shifting more towards high or very high risk scores (Fig. \ref{fig:commvulex-ne}). This shift  demonstrates that regional communities are increasing their dependence on species that are highly vulnerable to changing ocean conditions for their commercial revenue. Strategies for management should account for this increased reliance on climate-sensitive stocks.

```{r commvulex-ne, fig.width = 6.5, fig.asp=.5, fig.cap=return_caption(chunk_name = "commvulex", region = "NewEngland")}
return_plot("commvulex", region = "NewEngland")
```

### Implications
Social and demographic indicators highlight potential vulnerabilities in New England's most active commercial fishing ports. Industry participants in these locations face increased risk from shifting fishing patterns, whether driven by new regulations or broader ecosystem changes. Because many of these primary communities show medium to high socio-demographic risk, they may lack the necessary resources to adapt effectively to industry transitions

```{r, eval = params$id_child_docs, results = "asis"}
<<child-footer>>
```