```{r, eval = FALSE, results = "asis"}
<<child-header>>
```

```{r, eval = params$draft}
<<ma-draft>>
```

[As noted above](#climate-and-ecosystem-change), the Mid-Atlantic [Cold Pool](https://noaa-edab.github.io/catalog/cold_pool.html) is a summer to early fall feature that creates seasonally suitable habitat for some species. Cold pool persistence has decreased indicating that the duration of the cold pool habitat is shorter compared to the 1960s (Fig. \ref{fig:cold-pool-time}). However, all cold pool indices were near or above the long-term average in 2025 and likely related to the influx of northern waters into the system (see [2024 Highlights](https://noaa-edab.github.io/catalog/observation_synthesis_2024.html) ). A change in the timing of the autumn breakdown of the Cold Pool may impact the recruitment of species that rely on it for seasonal cues and habitat. Southern New England-Mid Atlantic yellowtail flounder recruitment and settlement are related to the strength of the MAB Cold Pool (a factor of extent and persistence). The correlation of pre-recruit settlers to the Cold Pool is thought to represent a bottleneck in yellowtail flounder life history, whereby a local and temporary increase in bottom temperature can negatively impact the survival of settlers. Including the effect of Cold Pool variations on yellowtail recruitment reduced retrospective patterns and improved predictive skill in a stock assessment model. This connection is especially important given the long-term decline in the duration of the Cold Pool.

```{r cold-pool-time, fig.cap=return_caption(chunk_name = "cold-pool-time", region = "BothReports"), fig.width = 4, fig.asp=.52}
return_plot("cold-pool-time", region = "BothReports")
```

The seasonal timing of Mid-Atlantic [phytoplankton](https://noaa-edab.github.io/catalog/chl_pp.html) blooms shows high interannual variability during the fall bloom period (October-December, Fig. \ref{fig:chl-month}). The significant increase in January chlorophyll suggests that the fall bloom period is continuing into the winter, with higher phytoplankton concentrations now than in the late 1990s. The significant decrease of chlorophyll in September could be related to warmer temperatures persisting into early fall and nutrient limitation causing a delay in the fall bloom. Changes to bloom timing can create a mismatch with the timing of larval fish development and may impact recruitment.

```{r chl-month, out.width = "100%", fig.cap=return_caption(chunk_name = "chl-month", region = "MidAtlantic")}
return_plot("chl-month", region = "MidAtlantic")
```

```{r, eval = FALSE, results = "asis"}
<<child-footer>>
```