```{r, eval = params$id_child_docs, results = "asis"}
<<child-header>>
```

```{r, eval = params$draft}
<<ma-draft>>
```

#### Indicators: Fish productivity and condition shifts
Indicators of [fish productivity](https://noaa-edab.github.io/catalog/productivity_anomaly.html) are derived from observations (surveys) or models (stock assessments). Fish productivity declined during the 1990’s and 2000’s with declining production of summer flounder and has been variable since, as described by the small-fish-per-large-fish anomaly indicator (derived from NEFSC bottom trawl survey) (Fig. \ref{fig:productivity-anomaly}). Bluefish, black sea bass, and goosefish have sporadic years with large positive anomalies, but most years have small negative anomalies. This decline in fish productivity is also shown by a similar analysis based on stock assessment model outputs (recruitment per spawning stock biomass anomaly). Most species had positive recruitment anomalies in the 1990s and 1990s and are currently showing negative anomalies, indicating a decline in productivity. Fish productivity can be affected by parental condition, environmental conditions, timing and availability of prey for recruits, as well as retention of recruits within favorable habitat. High offshore advection during spawning seasons can reduce recruitment success and affect overall fish productivity.

```{r productivity-anomaly, fig.cap = return_caption(chunk_name = "productivity-anomaly", region = "MidAtlantic"), fig.width=6.5, fig.asp=0.6}
return_plot("productivity-anomaly", region = "MidAtlantic")
```

The health of individual fish (i.e., fish condition, measured as weight for a given length) can contribute to population productivity through improved growth, reproduction and survival. Mid-Atlantic [fish condition](https://noaa-edab.github.io/catalog/condition.html)  was generally high to very high prior to 2000, low to very low from 2001-2010 (concurrent with declines in productivity, Fig. \ref{fig:productivity-anomaly})), and mixed since 2011. In 2025, condition continued to be mixed, with general improvement since a relatively low condition year in 2021 (Fig. \ref{fig:mab-cf}). Preliminary analyses show that years dominated by small copepods and warmer spring temperatures may improve fish condition for Atlantic mackerel and butterfish. Similar environmental drivers may be important to other species.

```{r mab-cf, fig.cap = return_caption(chunk_name = "mab-cf", region = "MidAtlantic"), fig.width=6.5, fig.asp=.7}
return_plot("mab-cf", region = "MidAtlantic")
```

#### Drivers:

`r risks_setting_catch_drivers`

##### Biological Drivers: Forage quality and abundance

`r forage_quality`

Forage fish [energy content](https://noaa-edab.github.io/catalog/energy_density.html) fluctuates based on growth, reproduction, environmental conditions, and ecosystem productivity. In the Mid-Atlantic Bight, butterfish are the most abundant high ED forage species (Fig. \\ref{fig:energy-density}), though their fall ED have recently declined toward lower spring averages. Atlantic herring and Atlantic mackerel also serve as high-energy prey, but herring show recently low ED and declining abundance, while mackerel are most abundant in the spring despite having higher ED in the fall. Moderate energy forage species (longfin squid, northern shortfin squid, and silver hake) are of intermediate abundance and show minimal annual and seasonal variation in ED. Other species have high ED but lower abundance decreasing their reliability as a food source.

<!-- MAY NEED TO REPLACE WITH STATIC FIGURE  -->
```{r energy-density, fig.cap=return_caption(chunk_name = "energy-density", region = "BothReports"),  fig.width = 6.5, fig.asp = 0.5}
return_plot("energy-density", region = "BothReports")
```

Changes in the overall abundance of forage fish can influence managed species productivity as it relates to changes in food availability. A spatially-explicit [forage index](https://noaa-edab.github.io/catalog/forage_index.html) for the Mid-Atlantic shows a long term declining trend in fall, with higher forage biomass in fall than spring (Fig. \ref{fig:foragebio}). Forage biomass was highest during fall in the early-1980s. The decrease of fall forage biomass in the Mid-Atlantic may reduce the health and reproductive output of fish species. Additionally, this may be exacerbated by lower energy densities of prey, especially in years of higher water temperatures when metabolic demands are higher. 

```{r foragebio, fig.cap = return_caption(chunk_name = "foragebio", region = "MidAtlantic"), fig.asp=.6}
return_plot("foragebio", region = "MidAtlantic")
```

[Benthic invertebrates](https://noaa-edab.github.io/catalog/benthos_index.html) are extremely important forage for some managed species (e.g., black sea bass). Macrobenthos are small benthic organisms that tend to be prey for larger benthos and benthivores. Macrobenthos indices show long-term declines in spring (Fig. \ref{fig:benthos}), indicating a potential decrease in food availability for their predators. In contrast, Mid-Atlantic megabenthos indices show long-term increases in spring. Fish productivity may be positively impacted in recent years for juvenile fish that target macrobenthos, such as small crustaceans and polychaetes, and negatively impacted for fish such as black sea bass and striped bass that target megabenthos such as crabs. Other species that are generalist feeders such scup and skates may not be as impacted by offsetting trends in the benthic community.

```{r benthos, fig.cap=return_caption(chunk_name = "benthos", region = "MidAtlantic"), fig.width=6.5, fig.asp=.4}
return_plot("benthos", region = "MidAtlantic")
```

##### Biological Drivers: Lower trophic levels

[Phytoplankton](https://noaa-edab.github.io/catalog/chl_pp.html) are the foundation of the food web and are the primary food source for zooplankton and filter feeders such as shellfish. Multiple environmental and oceanographic drivers affect the abundance, [composition](https://noaa-edab.github.io/catalog/chl_pp.html), spatial distribution, and productivity of phytoplankton. While changes in phytoplankton productivity could affect fish productivity (including the productivity of forage fish), there is no clear long-term trend in Mid-Atlantic total primary production (Fig. \ref{fig:totpp}).

Changing [zooplankton abundance](https://noaa-edab.github.io/catalog/zoo_abundance_anom.html) abundance may impact forage fish energy content and abundance, as well as the prey field of filter feeding whales, and managed species through food web impacts. Mid-Atlantic indices show high variability without a clear trend for large copepods, while small-bodied copepods (*Calanus finmarchicus*) show long-term and recent decreases, and krill (Euphausiids) show increasing trends (Fig. \ref{fig:zoopanom}).  Energy density varies by season and location, with high-energy large copepods most abundant on the Northeast shelf from April through June. The community is undergoing a systemic shift away from copepod dominance and toward increased krill presence.

```{r zoopanom, fig.cap=return_caption(chunk_name = "zoopanom", region = "MidAtlantic"), fig.width=6.5, fig.asp=.4}
return_plot("zoopanom", region = "MidAtlantic")
```

##### Environmental Drivers

Fish production can also be directly related to the prevailing environmental conditions by altering metabolism (growth), reproductive processes, and survival. Marine species possess thermal tolerances and can experience stressful or lethal conditions if water temperatures exceed certain levels. We have observed in past years extreme temperatures at both the [surface](https://noaa-edab.github.io/catalog/seasonal_oisst_anom.html) and [bottom](https://noaa-edab.github.io/catalog/bottom_temp_model_anom.html that exceed [thermal tolerance](https://noaa-edab.github.io/catalog/thermal_habitat_gridded.html) limits for some fish and shellfish. However, in 2025, Mid-Atlantic surface and bottom temperatures were near or below the long-term average and the amount of habitat exceeding a 24 oC thermal tolerance was limited to the southern MAB, where those conditions occurred for fewer than 30 days (Fig. \ref{fig:therm-hab-persist-2024}).

A single surface [marine heatwave](https://noaa-edab.github.io/catalog/heatwave.html) occurred in the Mid-Atlantic Bight in 2025, starting July 15th and lasting seven days. This brief event was the only heatwave recorded across the entire Continental Shelf for the year. The MAB experienced six surface marine cold spells in 2025, including an event in February that ranked as the 8th strongest on record. Additionally, a significant bottom cold spell occurred in January, lasting 57 days and ranking as the 5th strongest on record. During this period, bottom temperatures averaged 7.2 °C, nearly 2 °C lower than the historical average.

Lower ocean temperatures near long-term averages will affect species differently across the region. While cold-water species like cod may benefit from these conditions, warm-water species such as black sea bass are unlikely to see positive effects. This variability in regional cooling highlights the need for management to account for shifting species distributions and productivity.

```{r therm-hab-persist-2024, fig.cap = return_caption(chunk_name = "therm-hab-persist-2024", region = "BothReports"), fig.width=6.5, fig.asp=.5}
return_plot("therm-hab-persist-2024", region = "BothReports")
```

The newly-developed advection index (Fig. \ref{fig:advection-index}) shows total transport of water onto and off the continental shelf which can impact the survival of early life stages of fish and invertebrates. Long-term trends in the Mid-Atlantic show increased offshore movement of surface and bottom waters in June, which could decrease retention of some species. Further species level studies are needed to link spawning timing and larval periods to advection trends at the corresponding spatial and temporal scalesdepth and month.

```{r advection-index, fig.cap = return_caption(chunk_name = "advection-index", region = "BothReports"), fig.width=6.5, fig.asp=.5}
return_plot("advection-index", region = "MidAtlantic")
```

`r oa_paragraph`

```{r mab-oa, eval = TRUE, out.width = '100%', fig.cap = return_caption(chunk_name = "mab-oa", region = "BothReports")}
return_plot("mab-oa", region = "BothReports")
```

Low dissolved oxygen levels (< 5 mg/L) remained localized and brief on the [MAB shelf in 2025](https://noaa-edab.github.io/catalog/observation_synthesis_2023.html), resulting in no industry-reported mass mortality events despite the potential for hypoxia to reduce species growth or cause death. Localized hypoxia (< 2 mg/L) occurred nearshore east of Point Pleasant, NJ, southwest of Newport, RI, and at the western end of the Cape Cod Canal, while broader shelf-wide levels below 5 mg/L were not widespread. This contrast follows previous years where [hypoxic](https://noaa-edab.github.io/catalog/observation_synthesis_2023.html) events in Cape Cod Bay (2019, 2020) and off New Jersey (2023) potentially caused fish, lobster, and crab [mortality](https://sebsnjaesnews.rutgers.edu/2023/12/rutgers-scientists-observe-unusual-ocean-conditions-possibly-linked-to-mortality-in-marine-life-off-new-jersey/). While shelf-wide monitoring data is currently limited, biological and oceanographic drivers of oxygen levels continue to be tracked to assess the duration and extent of future events.

##### Drivers: Predation
The abundance and distribution of marine mammal, shark predators, and other Atlantic Highly Migratory Species (HMS), can affect both the productivity and mortality rates on managed stocks. Predators can consume managed species or compete for the same resources, resulting in increased natural mortality or decreased productivity. The northeast shift in [whales and dolphins](https://noaa-edab.github.io/catalog/cetacean_dist.html) (Fig. \ref{fig:protectedspp-dist-shifts}) indicates a change in the overlap between marine mammals and managed fishes. Since we also observe distribution shifts in managed species as well as forage species, the effect of changing predator distributions alone is difficult to quantify.

Indicators for shark populations, combined with information on gray seals (see [Protected Species Implications section, above](#protected-species)), suggests predator populations range from stable ([sharks](https://noaa-edab.github.io/catalog/hms_cpue.html)) to increasing ([gray seals](https://noaa-edab.github.io/catalog/seal_pups.html))  in the MAB. [Stock status](https://noaa-edab.github.io/catalog/hms_stock_status.html) is mixed for HMS stocks (including sharks, swordfish, billfish, and tunas) occurring throughout the Northeast U.S. shelf. While there are several HMS species considered to be overfished or that have unknown stock status, the population status for some managed Atlantic sharks and tunas is at or above the biomass target, suggesting the potential for robust (or rebuilt) predator populations and subsequent predation pressure on managed species. Increasing predator populations or changing distribution of predators may result in increased predation pressure.


```{r, eval = params$id_child_docs, results = "asis"}
<<child-footer>>
```
