```{r, eval = params$id_child_docs, results = "asis"}
<<child-header>>
```

```{r, eval = params$draft}
<<ne-draft>>
```

#### Indicators: Fish productivity and condition shifts

Indicators of [fish productivity](https://noaa-edab.github.io/catalog/productivity_anomaly.html) are derived from observations (surveys) or models (stock assessments). With the exception of two years (2006 and 2013), fish productivity has been below the long-term average in the Gulf of Maine since the early 2000s, as described by the small-fish-per-large-fish anomaly indicator (derived from NEFSC bottom trawl survey)(Fig. \ref{fig:productivity-anomaly}). This decline in fish productivity is also shown by a similar analysis based on stock assessment model outputs (recruitment per spawning stock biomass anomaly). Other signs of changing productivity in New England are the declines in [common tern chicks](https://noaa-edab.github.io/catalog/seabird_ne.html) per nest (Fig. \ref{fig:seabird-ne-productivity}) and continued low returns of hatchery [Atlantic salmon](https://noaa-edab.github.io/catalog/gom_salmon.html)(Fig. \ref{fig:salmon}) despite short-term increases in adult salmon numbers.
```{r productivity-anomaly, fig.cap = return_caption(chunk_name = "productivity-anomaly"), fig.width=6.5, fig.asp=1}
productivity_anomaly_plot / recruit_anomaly_plot
```

```{r seabird-ne-productivity, fig.cap = return_caption(chunk_name = "seabird-ne-productivity"),  fig.width=6.5, fig.asp=.3}
return_plot("seabird-ne-productivity")
```

```{r salmon, fig.cap = return_caption(chunk_name = "salmon"),fig.width = 6.5,fig.asp = 0.3}
return_plot("salmon")
```
The health of individual fish (i.e., fish condition) can contribute to population productivity through improved growth, reproduction, and survival. [Fish condition](https://noaa-edab.github.io/catalog/condition.html) in the Gulf of Maine and Georges Bank regions were generally good prior to 2000, poor from 2001-2010 (concurrent with declines in fish productivity, Fig. \ref{fig:productivity-anomaly}), and a mix of good and poor since 2011. In 2024, fish condition was poor for most species on both Georges Bank and in the Gulf of Maine (Fig. \ref{fig:ne-cf}).Preliminary analyses show that changes in temperature, zooplankton, fishing pressure, and population size influence the condition of different fish species.
```{r ne-cf, fig.cap = return_caption(chunk_name = "salmon"), fig.width=6.5, fig.asp=1.2}
return_plot("ne-cf")
```

#### Drivers 

Fish productivity and condition are affected by increasing metabolic demands from increasing temperature, combined with changes in the availability and quality of prey. Long-term environmental trends and episodic extreme temperatures, ocean acidification, and low oxygen events represent multiple stressors that can affect growth rates, reproductive success and recruitment, and cause mortality.

##### Biological Drivers: Forage quality and abundance

The amount of forage fish available in the ecosystem combined with the energy content of the forage species determines the amount of energy potentially available to predators in the ecosystem. Changes in the forage fish base can drive managed and protected species production and condition.

The [energy content](https://noaa-edab.github.io/catalog/energy_density.html) of juvenile and adult forage fish as prey is related to forage fish growth and reproductive cycles, as well as environmental conditions. The energy content of Atlantic herring was estimated to be highest of any forage species in the 1980s and 1990s, based on very small numbers of fish. Most observations from the NEFSC trawl surveys are below the previous estimates (Fig. \ref{fig:energy-density}). However, a recent study that included samples from additional sources indicated herring energy density peaked in summer, with some values closer to the historic estimates. Silver hake, longfin squid (Loligo in figure), and shortfin squid (Illex in figure) remain lower than previous estimates.

```{r energy-density, fig.cap=return_caption(chunk_name = "energy-density"),  fig.width = 6.5, fig.asp = 0.5}
return_plot("energy-density")
```
Changes in the overall abundance of forage fish can influence managed species productivity as it relates to changes in food availability. New England [fall forage biomass](https://noaa-edab.github.io/catalog/forage_index.html) is stable with long-term increases in the spring GOM (Fig. \ref{fig:foragebio}). Forage biomass was highest during fall in the 1980s.

```{r foragebio, fig.cap = return_caption(chunk_name = "foragebio"), fig.width = 6.5, fig.asp=.3}
return_plot("foragebio")
```

[Benthic invertebrates](https://noaa-edab.github.io/catalog/benthos_index.html) are extremely important forage for some managed species (e.g. flatfish, juvenile cod and haddock) Macrobenthos indices show long term declines in spring. In contrast, megabenthos indices show long-term increases during the fall in both GB and GOM (Fig. \ref{fig:benthos}).

```{r benthos, fig.cap=return_caption(chunk_name = "benthos"), fig.width=6.5, fig.asp=.9}
return_plot("benthos")
```

##### Biological Drivers: Lower trophic levels

Phytoplankton are the foundation of the marine food web and are the primary food source for zooplankton and filter feeders such as shellfish. Multiple environmental and oceanographic drivers affect the abundance, composition, spatial distribution, and productivity of phytoplankton. While changes in phytoplankton productivity could affect fish productivity (including forage), there is no clear long-term trend in New England total primary production (Fig. \ref{fig:totpp}).

[Zooplankton communities](https://noaa-edab.github.io/catalog/zoo_abundance_anom.html)  in New England have been changing in composition. There has been an increase in GOM small bodied copepods, a recent decline in large bodied copepods in GB, and long-term increases in euphausids in GB and GOM (Fig. \ref{fig:zoopanom}). A changing mix of zooplankton prey can impact forage fish energy content and abundance, as well as the prey field of filter feeding whales.

Since 2010, the abundance of the lipid-rich older stages of  [*Calanus finmarchicus*](https://noaa-edab.github.io/catalog/wbts_mesozooplankton.html) in the GOM has declined. Observations from a fixed time series station in Wilkinson Basin  indicate that *Calanus* [seasonal abundance](https://noaa-edab.github.io/catalog/calanus_variation.html) in late summer-winter between 2020-2022 has declined to 20-40% of its population level in 2005-2008 but has rebounded to 60-70% in 2024 due to the presence of colder water(Fig. \ref{fig:zooplankton-season}). However, spring abundances are still the same as 15-20 years ago. The seasonal differences in abundance change reflect differences in influence of primary seasonal drivers:

1. *Calanus* reproductive output is tied to phytoplankton availability in late winter/early spring.
2. Gulf of Maine source waters drive *Calanus* supply (high *Calanus* in Scotian Shelf/Labrodor shelf water (LSW) and less in warm slope water (WSW))
3. Predation is likely higher with warmer temperatures


```{r zoopanom, fig.cap=return_caption(chunk_name = "zoopanom"), fig.width=6.5,fig.asp = 1}
return_plot("zoopanom")
```

```{r zooplankton-season, fig.cap=return_caption(chunk_name = "zooplankton-season")}
return_plot("zooplankton-season")
```

##### Environmental Drivers

Fish production can also be directly related to the prevailing environmental conditions by altering metabolic (growth) and reproductive processes. Many species possess thermal tolerances and can experience stressful or lethal conditions if temperatures exceed certain levels. Extreme temperature at both the [surface](https://noaa-edab.github.io/catalog/seasonal_oisst_anom.html) (Fig. \ref{fig:long-term-sst}) and [bottom](https://noaa-edab.github.io/catalog/bottom_temp_model_anom.html) can exceed [thermal tolerance](https://noaa-edab.github.io/catalog/thermal_habitat_gridded.html) limits for some fish. For example, 2012 had among the warmest surface and bottom temperatures (GB) in New England. A large proportion of the Georges Bank and Mid-Atlantic regions had bottom temperatures above the 15℃ thermal tolerance for most groundfish, with some days in the Mid-Atlantic exceeding the 24℃ potential mortality limit (Fig. \ref{fig:therm-hab-persist-2024}). 

In 2024, only one [surface marine heatwave](https://noaa-edab.github.io/catalog/heatwave_year.html) occurred throughout the entire U.S. Northeast Shelf due to the cooler ocean conditions observed in the region.  This surface marine heatwave occurred in the Gulf of Maine starting on May 29th, peaking on June 7th, and lasting 12 days.  This marine heatwave was not within the top 10 on record in terms of intesity.
```{r therm-hab-persist-2024, fig.cap=return_caption(chunk_name = "therm-hab-persist-2024"), fig.width=6.5, fig.asp=0.5}
return_plot("therm-hab-persist-2024")
```

`r oa_paragraph`

```{r mab-oa, eval = TRUE, out.width = '100%', fig.cap = return_caption(chunk_name = "mab-oa")}
return_plot("mab-oa")
```

Biological and oceanographic processes can affect the amount of oxygen present in the water column. During low oxygen (hypoxic) events, species' growth is negatively affected and very low oxygen can result in mortality. The duration and extent of hypoxic events is being monitored, but long-term shelf-wide observations are not yet available. However, [hypoxic events](https://noaa-edab.github.io/catalog/observation_synthesis_2023.html) were detected off the coast of New Jersey in 2023 and were potentially responsible for fish, lobster, and crab [mortalities](https://sebsnjaesnews.rutgers.edu/2023/12/rutgers-scientists-observe-unusual-ocean-conditions-possibly-linked-to-mortality-in-marine-life-off-new-jersey/). No hypoxic events were observed on the NE shelf in 2024.

##### Drivers: Predation

The abundance and distribution of predators can affect both the productivity and mortality rates on managed stocks. Predators can consume managed species or compete for the same resources resulting in increased natural mortality or declining productivity, respectively. The northeast shift in some [highly migratory species](https://noaa-edab.github.io/catalog/cetacean_dist.html) (Fig. \ref{fig:protectedspp-dist-shifts}) indicates a change in the overlap between predators and prey. Since we also observe distribution shifts in both managed and forage species, the effect of changing predator distributions alone is difficult to quantify.

[Gray seals](https://noaa-edab.github.io/catalog/grayseal.html) are fish predators with increasing populations in New England, however they are broad generalist feeders that do not generally target commercially-sized managed species. [Stock status](https://noaa-edab.github.io/catalog/hms_stock_status.html) is mixed for Atlantic Highly Migratory Species (HMS) stocks (including sharks, swordfish, billfish, and tunas) occurring throughout the Northeast U.S. shelf. While there are several HMS species considered to be overfished or that have unknown stock status, the population status for some managed Atlantic sharks and tunas is at or above the biomass target, suggesting the potential for robust (or rebuilt) predator populations among these managed species. Stable predator populations suggest stable predation pressure on managed species, but increasing predator populations may reflect increasing predation pressure.

#### Future Considerations

The processes that control fish productivity and mortality are dynamic, complex, and the result of the interactions between multiple system drivers. There is a real risk that short-term predictions in assessments and rebuilding plans that assume unchanging underlying conditions will not be as effective, given the observed ecological and environmental process changes documented throughout the report. Assumptions for species’ growth, reproduction, and natural mortality should continue to be evaluated for individual species. With observations of system-wide productivity shifts of multiple managed stocks, more research is needed to determine whether regime shifts or ecosystem reorganization are occurring, and how this should be incorporated into management

```{r, eval = params$id_child_docs, results = "asis"}
<<child-header>>
```