#### Indicators: Fish productivity and condition shifts
Indicators of [fish productivity](https://noaa-edab.github.io/catalog/productivity_anomaly.html) are derived from observations (surveys) or models (stock assessments). Fish productivity has been declining in the Mid-Atlantic since the early 2000s, as described by the small-fish-per-large-fish anomaly indicator (derived from NEFSC bottom trawl survey) (Fig. \ref{fig:productivity-anomaly}). This decline in fish productivity is also shown by a similar analysis based on stock assessment model outputs (recruitment per spawning stock biomass anomaly).  

```{r productivity-anomaly, fig.cap = return_caption(chunk_name = "productivity-anomaly"), fig.width=6.5, fig.asp=0.6}
```

The health of individual fish (i.e. fish condition, measured as weight for a given length) can contribute to population productivity through improved growth, reproduction and survival. [Fish condition](https://noaa-edab.github.io/catalog/condition.html) in the MAB was generally good prior to 2000, poor from 2001-2010 (concurrent with declines in productivity, Fig. \ref{fig:productivity-anomaly}), and a mix of good and poor since 2011. In 2024, condition continued to be mixed, with general improvement since a relatively low condition year in 2021 (Fig. \ref{fig:mab-cf}). Preliminary analyses show that changes in temperature, zooplankton, fishing pressure, and population size influence the condition of different fish species.

```{r mab-cf, fig.cap = return_caption(chunk_name = "mab-cf"), fig.width=6.5, fig.asp=.7}
```

#### Drivers:

`r risks_setting_catch_drivers`

##### Biological Drivers: Forage quality and abundance

`r forage_quality`

```{r energy-density, fig.cap=return_caption(chunk_name = "energy-density"),  fig.width = 6.5, fig.asp = 0.5}
```

Changes in the overall abundance of forage fish can influence managed species productivity as it relates to changes in food availability. A spatially-explicit [forage index](https://noaa-edab.github.io/catalog/forage_index.html) for the Mid-Atlantic shows a long term declining trend in fall, with higher forage biomass in fall than spring (Fig. \ref{fig:foragebio}). Forage biomass was highest during fall in the early-1980s.   

```{r foragebio, fig.cap = return_caption(chunk_name = "foragebio"), fig.asp=.6}
```

[Benthic invertebrates](https://noaa-edab.github.io/catalog/benthos_index.html) are extremely important forage for some managed species (black sea bass, yellowtail and winter flounders, Atlantic cod, and haddock, and many skate species). Macrobenthos indices show long term declines in spring (Fig. \ref{fig:benthos}). In contrast, megabenthos indices show long-term increases in spring in the MAB.

```{r benthos, fig.cap=return_caption(chunk_name = "benthos"), fig.width=6.5, fig.asp=.4}
```

##### Biological Drivers: Lower trophic levels

[Phytoplankton](https://noaa-edab.github.io/catalog/chl_pp.html) are the foundation of the food web and are the primary food source for zooplankton and filter feeders such as shellfish. Multiple environmental and oceanographic drivers affect the abundance, [composition](https://noaa-edab.github.io/catalog/chl_pp.html), spatial distribution, and productivity of phytoplankton. While changes in phytoplankton productivity could affect fish productivity (including forage), there is no clear long-term trend in Mid-Atlantic total primary production (Fig. \ref{fig:totpp}).

[Zooplankton](https://noaa-edab.github.io/catalog/zoo_abundance_anom.html) communities in the Mid-Atlantic have high variability without trend for large copepods (Fig. \ref{fig:zoopanom}), long term and recent decreasing trends for smaller bodied copepods, and increasing trends for krill (Euphausiids). Small copepods are less energy-rich than Euphausiids (krill) or the larger-bodied copepod *Calanus finmarchicus*. A changing mix of zooplankton prey can impact forage fish energy content and abundance, as well as the prey field of filter feeding whales. 

```{r zoopanom, fig.cap=return_caption(chunk_name = "zoopanom"), fig.width=6.5, fig.asp=.4}
```

##### Environmental Drivers

Fish production can also be directly related to the prevailing environmental conditions by altering metabolic (growth) and reproductive processes. Many species possess thermal tolerances and can experience stressful or lethal conditions if temperatures exceed certain levels. Extreme temperatures at both the [surface](https://noaa-edab.github.io/catalog/seasonal_oisst_anom.html) and [bottom](https://noaa-edab.github.io/catalog/bottom_temp_model_anom.html) can exceed [thermal tolerance](https://noaa-edab.github.io/catalog/thermal_habitat_gridded.html) limits for some fish, which we have observed in past years. However, in 2024, surface and bottom temperatures were near or below normal in the MAB. The amount of habitat exceeding a 24 C thermal tolerance was limited to the southern MAB and the conditions persisted for fewer than 30 days (Fig. \ref{fig:therm-hab-persist-2024}).

In 2024, there were no [marine heatwaves](https://noaa-edab.github.io/catalog/heatwave.html) in the MAB. Only one marine heatwave occurred throughout the entire U.S. Northeast Shelf due to the cooler ocean conditions observed in the region.  This surface marine heatwave occurred in the [Gulf of Maine](https://noaa-edab.github.io/catalog/heatwave_year.html#newengland-22) starting on May 29th, peaking on June 7th, and lasting 12 days.

```{r therm-hab-persist-2024, fig.cap = return_caption(chunk_name = "therm-hab-persist-2024"), fig.width=6.5, fig.asp=.5}
```

`r oa_paragraph`

```{r mab-oa, eval = TRUE, out.width = '100%', fig.cap = return_caption(chunk_name = "mab-oa")}
```

Low oxygen was detected on the [MAB shelf in 2023](https://noaa-edab.github.io/catalog/observation_synthesis_2023.html), but not in 2024. Biological and oceanographic processes can affect the amount of oxygen present in the water column. During low oxygen (hypoxic) events, species growth is negatively affected, and very low oxygen can result in [mortality](https://sebsnjaesnews.rutgers.edu/2023/12/rutgers-scientists-observe-unusual-ocean-conditions-possibly-linked-to-mortality-in-marine-life-off-new-jersey/).

##### Drivers: Predation
The abundance and distribution of marine mammal, HMS, and shark predators can affect both the productivity and mortality rates on managed stocks. Predators can consume managed species or compete for the same resources, resulting in increased natural mortality or decreased productivity. The northeast shift in [whales and dolphins](https://noaa-edab.github.io/catalog/cetacean_dist.html)  (Fig. \ref{fig:protectedspp-dist-shifts}) indicates a change in the overlap between predators and prey. Since we also observe distribution shifts in managed species as well as forage species, the effect of changing predator distributions alone is difficult to quantify. 

Indicators for shark populations, combined with information on gray seals (see [Protected Species Implications section, above](#protected-species)), suggests predator populations range from stable ([sharks](https://noaa-edab.github.io/catalog/hms_cpue.html)) to increasing ([gray seals](https://noaa-edab.github.io/catalog/seal_pups.html))  in the MAB. [Stock status](https://noaa-edab.github.io/catalog/hms_stock_status.html) is mixed for Atlantic Highly Migratory Species (HMS) stocks (including sharks, swordfish, billfish, and tunas) occurring throughout the Northeast U.S. shelf. While there are several HMS species considered to be overfished or that have unknown stock status, the population status for some managed Atlantic sharks and tunas is at or above the biomass target, suggesting the potential for robust (or rebuilt) predator populations among these managed species. Stable predator populations suggest stable predation pressure on managed species, but increasing predator populations may reflect increasing predation pressure.

