```{r, eval = params$id_child_docs, results = "asis"}
<<child-header>>
```

```{r, eval = params$draft}
<<ma-draft>>
```

Based on federal vessel logbook data, [commercial fishery revenue](https://noaa-edab.github.io/catalog/wind_revenue.html) from trips in the current offshore wind lease areas, including the newly designated lease areas in the Central Atlantic, have varied annually from 2008-2023, with less than \$1 million in maximum annual revenue overlapping with these areas for most fisheries with the exception of the surfclam, monkfish, and longfin squid fisheries. Some fisheries see periodic spikes in revenue overlap with wind energy lease areas, including the surfclam (\$6.5 million), longfin squid (\$4.8 million), monkfish (\$2.5 million), and summer flounder (\$1.3 million) fisheries (Fig. \ref{fig:wea-spp-rev}). 

```{r wea-spp-rev, fig.cap=return_caption(chunk_name = "wea-spp-rev", region = "MidAtlantic"),  fig.width=5, fig.asp=.4}
return_plot("wea-spp-rev", region = "MidAtlantic")
```

Within Active Projects, the MAFMC managed fisheries most affected based on historic landings include longfin squid, monkfish, scup, Atlantic mackerel, and summer flounder, with a maximum of 6% of annual regional fishery revenue for chub mackerel occurring within Active Projects during 2008-2024, 5% for bluefish, 4% for butterfish, and 3% each for monkfish, scup, black sea bass, and longfin squid, respectively (see Table \ref{tab:wea-landings-rev}). Future offshore wind development may increase effects on these and additional species if more projects begin construction. Future fishery resource overlap with wind leases, especially surfclams and ocean quahogs, may change due to species distribution shifts attributable to climate change and recruitment and larval dispersion pattern changes caused by hydrodynamic flow disruptions from turbine foundations, which could also affect fishery landings/revenue.

```{r wea-landings-rev, ft.arraystretch = 1}
create_wea_landings_rev_table(council = "MAFMC") |>
  flextable::flextable() %>%
  flextable::set_caption("Mid-Atlantic managed species Landings and Revenue from Wind Energy Areas. *Less than a maximum of 50,000 lb was reported landed annually in wind energy lease areas for these species.") %>%
  flextable::width(width = c(2,2,2))
```

`r wind_interactions`

`r socio_intro`

```{r wea-port-rev, fig.cap=return_caption(chunk_name = "wea-port-rev", region = "MidAtlantic"),  fig.width=6.5, fig.asp=.82, fig.align='left'}
return_plot("wea-port-rev", region = "MidAtlantic")
```

Based on federal vessel logbook data, Point Lookout, NY (5.5% average, 17% maximum) and Virginia Beach, VA (3% average, 7.5% maximum) have the highest potential revenue loss from the Active Projects based on 2008-2024 total port fisheries revenue. Fewer Mid-Atlantic ports are affected by the Active Projects to date, as most are in the southern New England region, with the exception of CVOW and Empire Wind 1 (Fig. \\ref{fig:wea-port-rev}). Additional fishing revenue may be lost as more areas historically used for fishing are developed for offshore wind energy. In seven New England ports, Mid-Atlantic managed species account for at least 50% of landings from the Active Project areas by value or weight (Fig. \ref{fig:wind-rev}). Furthermore, impacts of offshore wind development may unevenly affect individual operators, with some permit holders deriving a much higher proportion of revenue from wind areas than the port-based mean.

```{r wind-rev, fig.cap=return_caption(chunk_name = "wind-rev", region = "MidAtlantic"), fig.width=6.5, fig.asp=.35, fig.align='left'}
return_plot("wind-rev", region = "MidAtlantic")
```

Top fishing communities with [socio-demographic concerns](https://noaa-edab.github.io/catalog/engagement.html) (i.e., Atlantic City, NJ,  and Hampton Bays, NY) should be considered in decision making to reduce the social and economic impacts and aid in the resilience and adaptive capacity of underserved communities. These are communities where we need to provide further resources to reach underserved and underrepresented groups and create opportunities for and directly involve these groups in the decision-making process.

```{r, eval = params$id_child_docs, results = "asis"}
<<child-footer>>
```
