```{r, eval = params$id_child_docs, results = "asis"}
<<child-header>>
```

```{r, eval = params$draft}
<<ne-draft>>
```

Offshore wind indicators are based on federal logbook data and do not include all data for all fisheries; therefore a complete evaluation of potential offshore wind energy development impacts would need to be supplemented by other data sources. For further information on the utility of the data, see the [socioeconomic impacts of offshore wind development data reports page](https://www.fisheries.noaa.gov/resource/data/socioeconomic-impacts-atlantic-offshore-wind-development).

Based on federal vessel logbook data, [commercial fishery revenue](https://noaa-edab.github.io/catalog/wind_revenue.html) from trips within Active Projects varied annually from 2008-2024. Maximum annual revenue for the fisheries with the most overlap with Active Projects peaked at over \$8.7 million for the sea scallop fishery, \$1.1 million for monkfish, \$477,000 for skates, \$377,000 for yellowtail flounder, and $344,000 for Atlantic herring (Fig. \ref{fig:wea-spp-rev-ne}). Individual groundfish species are more affected on a percentage basis, with up to 13% of historical annual revenues overlapping with Active Projects for species such as little skate (13%), barndoor skate (11%), yellowtail flounder (10%), and 6% each for red hake, clearnose skate, and winter skate, respectively (Table \ref{tab:wea-landings-rev-ne}). Future fishery resource overlap with wind leases, especially scallops, may change due to species distribution shifts attributable to climate change and recruitment and larval dispersion pattern changes caused by hydrodynamic flow disruptions from turbine foundations, which could also affect fishery landings/revenue.

```{r wea-spp-rev-ne, fig.cap=return_caption(chunk_name = "wea-spp-rev", region = "NewEngland"),  fig.width=6, fig.asp=.6}
return_plot("wea-spp-rev", region = "NewEngland")
```

```{r wea-landings-rev-ne, ft.arraystretch = 1}
spp.mark = c('Barndoor Skate','Bluefish','Smooth Skate','Windowpane Flounder','Atlantic Cod','Conch','Offshore Hake','Cobia','Haddock','Atlantic Croaker','Witch Flounder')

create_wea_landings_rev_table(council = "NEFMC") |>
  dplyr::arrange(desc(`Maximum Percent Total Annual Regional Species Revenue`)) |> 
  dplyr::slice_head(n = 10) |>
  dplyr::mutate(
    `NEFMC, MAFMC, and ASMFC Managed Species` = dplyr::case_when(
      `NEFMC, MAFMC, and ASMFC Managed Species` %in% spp.mark ~ paste0(`NEFMC, MAFMC, and ASMFC Managed Species`, "*"),
      TRUE ~ `NEFMC, MAFMC, and ASMFC Managed Species`
    )
  ) |>
  flextable::flextable() %>%
  flextable::set_caption("Top 10 New England managed species Landings and Revenue from Active Projects. *Less than a maximum of 50,000 lb was reported landed annually in wind energy lease areas for these species.") %>%
  flextable::width(width = c(2,2,2))
```

`r socio_intro`

```{r wea-port-rev, fig.cap=return_caption(chunk_name = "wea-port-rev", region = "NewEngland"),  fig.width=6.5, fig.asp=.6, fig.align='left'}
return_plot("wea-port-rev", region = "NewEngland")
```

<!-- \newpage -->

Based on federal vessel logbook data, Little Compton, RI (13% average and 24% maximum) and Westport, MA (6% average and 12% maximum) have the highest potential revenue loss from Active Projects based on 2008-2024 total port fisheries revenue, with all other New England communities having less than 5% (Fig. \ref{fig:wea-port-rev}). Additional fishing revenue may be affected as more areas historically used for fishing are developed for offshore wind energy. In response to Council request, we also [present information](https://noaa-edab.github.io/catalog/wind_port.html) on the Mid-Atlantic ports that land NEFMC-managed species within the wind lease areas. This information can help fisheries managers better understand how NEFMC-managed species may be more broadly affected by offshore wind development. Seven Mid-Atlantic ports attribute the majority of their landings within the wind lease areas to NEFMC-managed species (Fig. \ref{fig:wind-rev}); note that overall, the Active Projects account for less than 17% of total revenue in these ports. Future analyses should also consider that the impacts of offshore wind development may unevenly affect individual operators, with some permit holders deriving a much higher proportion of revenue from wind areas than the port-based mean.

```{r wind-rev, fig.cap=return_caption(chunk_name = "wind-rev", region = "NewEngland"), fig.width=6.5, fig.asp=.35, fig.align='left'}
return_plot("wind-rev", region = "NewEngland")
```


BOEM reports that cumulative offshore wind development (if all proposed projects are developed) could have moderate impacts on low-income members of vulnerable communities who work in the commercial fishing and for-hire fishing industry due to disruptions to fish populations, restrictions on navigation, and increased vessel traffic as well as exacerbation of existing vulnerabilities of low-income workers to economic impacts.

Top fishing communities with high [socio-demographic](https://noaa-edab.github.io/catalog/engagement.html) and/or gentrification concerns such as Little Compton, RI, New Bedford, MA and New London, CT should be recognized as having potential additional vulnerability of the Active Projects and considered in decision making to reduce the social and economic impacts and aid in the resilience and adaptive capacity of these communities. In addition to fisheries landing overlaps,  New Bedford, MA and New London, CT also support significant offshore wind port infrastructure needs for the Active Projects. Historically, the introduction of new industries can trigger industrial and socioeconomic gentrification of fishing ports. Competition for port space and potential pivoting of space use for offshore wind development should be monitored closely to ensure fishing communities are not adversely impacted. Additionally, offshore wind could increase recreational fishing opportunities at the turbines, potentially creating a demand for additional tourism, recreational fishing and boating port space in communities already balancing these uses with commercial fishing infrastructure, for example Point Judith, RI, and Newport, RI, and Gloucester, MA. Socio-demographic concerns also highlight communities where further resources are needed to reach underserved and underrepresented groups and create opportunities for, and directly involve, these groups in the decision-making process.

```{r, eval = params$id_child_docs, results = "asis"}
<<child-footer>>
```
