```{r, eval = FALSE, results = "asis"}
<<child-header>>
```

```{r, eval = params$draft}
<<ma-draft>>
```

Offshore wind indicators are based on federal logbook data and do not include all data for all fisheries; therefore a complete evaluation of potential offshore wind energy development impacts would need to be supplemented by other data sources. For further information on the utility of the data, see the [socioeconomic impacts of offshore wind development data reports page](https://www.fisheries.noaa.gov/resource/data/socioeconomic-impacts-atlantic-offshore-wind-development).

[Commercial fishery revenue](https://noaa-edab.github.io/catalog/wind_revenue.html) from trips within Active Projects have varied annually from 2008-2024, with less than \$500,000 in maximum annual revenue overlapping with these areas for most fisheries with the exception of longfin squid (\$2 million), monkfish (\$1.1 million), ocean quahog (\$783,000), and summer flounder (\$556,000) in specific years  (Fig. \ref{fig:wea-spp-rev}). 

```{r wea-spp-rev, fig.cap=return_caption(chunk_name = "wea-spp-rev", region = "MidAtlantic"),  fig.width=5, fig.asp=.4}
return_plot("wea-spp-rev", region = "MidAtlantic")
```

Based on revenue between 2008-2024, MAFMC-managed fisheries derived a maximum of 6% of annual regional fishery revenue from the Active Project areas. The specific fisheries affected include chub mackerel (6% of revenue), bluefish (5%), butterfish (4%), and monkfish, scup, black sea bass, and longfin squid (3% each) (Table \ref{tab:wea-landings-rev}). Future offshore wind development may increase effects on these and additional species if more projects begin construction. Future fishery resource overlap with wind leases, especially surfclams and ocean quahogs, may change due to species distribution shifts attributable to climate change and recruitment and larval dispersion pattern changes caused by hydrodynamic flow disruptions from turbine foundations, which could also affect fishery landings and/or revenue.

<!-- \newpage -->

```{r wea-landings-rev, ft.arraystretch = 1}
spp.mark = c('Chub Mackerel','Bluefish','Golden Tilefish')

create_wea_landings_rev_table(council = "MAFMC") |>
  dplyr::arrange(desc(`Maximum Percent Total Annual Regional Species Revenue`)) |> 
  dplyr::slice_head(n = 10) |>
  dplyr::mutate(
    `NEFMC, MAFMC, and ASMFC Managed Species` = dplyr::case_when(
      `NEFMC, MAFMC, and ASMFC Managed Species` %in% spp.mark ~ paste0(`NEFMC, MAFMC, and ASMFC Managed Species`, "*"),
      TRUE ~ `NEFMC, MAFMC, and ASMFC Managed Species`
    )
  ) |>
  flextable::flextable() %>%
  flextable::set_caption("Top 10 Mid-Atlantic managed species landings and revenue from Active Projects. *These species did not have more than 50,000 lbs landed from the Active Project areas in any year.") %>%
  flextable::width(width = c(2,2,2))
```

`r socio_intro`

```{r wea-port-rev, fig.cap=return_caption(chunk_name = "wea-port-rev", region = "MidAtlantic"),  fig.width=6.5, fig.asp=.82, fig.align='left'}
return_plot("wea-port-rev", region = "MidAtlantic")
```

Based on federal vessel logbook data, Point Lookout, NY (5.5% average, 17% maximum) and Virginia Beach, VA (3% average, 7.5% maximum) have the highest potential revenue loss from the Active Projects based on 2008-2024 total port fisheries revenue. Fewer Mid-Atlantic ports are affected by the Active Projects to date, as most are in the southern New England region, with the exception of CVOW and Empire Wind 1 (Fig. \ref{fig:wea-port-rev}). Additional fishing revenue may be affected as more areas historically used for fishing are developed for offshore wind energy. 

In response to Council request, we also [present information](https://noaa-edab.github.io/catalog/wind_port.html) on the New England ports that land MAFMC-managed species within the wind lease areas. This information can help fisheries managers better understand how MAFMC-managed species may be more broadly affected by offshore wind development. Seven New England ports attribute the majority of their landings within the wind lease areas to MAFMC-managed species (Fig. \ref{fig:wind-rev}); note that two of these ports do not have a majority of MAFMC-managed landings when only the Active Project areas are considered. Additionally, the Active Projects account for less than 8% of total revenue in these seven ports. Future analyses should also consider that the impacts of offshore wind development may unevenly affect individual operators, with some permit holders deriving a much higher proportion of revenue from wind areas than the port-based mean.

```{r wind-rev, fig.cap=return_caption(chunk_name = "wind-rev", region = "MidAtlantic"), fig.width=6.5, fig.asp=.35, fig.align='left'}
return_plot("wind-rev", region = "MidAtlantic")
```

If all proposed projects are developed, BOEM reports that cumulative offshore wind development  could have moderate impacts on low-income members of communities who work in the commercial fishing and for-hire fishing industry due to disruptions to fish populations, restrictions on navigation and increased vessel traffic, and exacerbation of existing vulnerabilities of low-income workers to economic impacts.

Top fishing communities with both landings from the Active Projects (i.e., Point Lookout, NY and Newport News, VA) and [socio-demographic or gentrification concerns](https://noaa-edab.github.io/catalog/engagement.html) should be recognized as having additional vulnerability from Active Projects. To reduce further social and economic impacts and aid in the resilience and adaptive capacity of these communities, this vulnerability should be considered in decision making. Historically, the introduction of new industries can trigger industrial and socioeconomic gentrification of fishing ports. Competition for port space and potential pivoting of space use for offshore wind development should be monitored closely to ensure fishing communities are not adversely impacted. Additionally, offshore wind could increase recreational fishing opportunities at the turbines, potentially creating a demand for additional tourism, recreational fishing and boating port space in communities already balancing these uses with commercial fishing infrastructure (e.g., Virginia Beach, VA, Montauk, NY and Barnegat Light, NJ). Socio-demographic concerns also highlight communities where further resources are needed to reach underserved and underrepresented groups and create opportunities for, and directly involve, these groups in the decision-making process.

```{r, eval = FALSE, results = "asis"}
<<child-footer>>
```
