```{r, eval = params$id_child_docs, results = "asis"}
<<child-header>>
```

```{r, eval = params$draft}
<<ma-draft>>
```

`r wind_impact` in lease areas could be forgone or reduced and associated effort displaced when the Active Projects are developed. Commercial fishery revenue from trips within Active Projects have varied annually from 2008-2024, with less than \$500,000 in maximum annual revenue overlapping with these areas for most fisheries with the exception of longfin squid (\$2 million), monkfish (\$1.1 million), ocean quahog (\$783,000), and summer flounder (\$556,000) in specific years (Fig. \ref{fig:wea-spp-rev}). By percent of revenue affected,  specific fisheries include chub mackerel (6% of revenue), bluefish (5%), butterfish (4%), and monkfish, scup, black sea bass, and longfin squid (3% each) (Table \ref{tab:wea-landings-rev}). Displaced fishing effort can alter fishing area, timing, and method patterns, which can in turn change habitat, species (managed and protected), and fleet interactions. Several factors, including fishery regulations, fishery availability, and user conflicts affect where, when, and how fishing effort may be displaced, along with impacts to and responses of affected fish species. Future fishery resource overlap with wind leases, especially surfclams and ocean quahogs, may change due to species distribution shifts attributable to climate change and recruitment and larval dispersion pattern changes caused by hydrodynamic flow disruptions from turbine foundations, which could also affect fishery landings and/or revenue.

```{r wea-spp-rev, fig.cap=return_caption(chunk_name = "wea-spp-rev", region = "MidAtlantic"),  fig.width=5, fig.asp=.4}
return_plot("wea-spp-rev", region = "MidAtlantic")
```

Point Lookout, NY (5.5% average, 17% maximum) and Virginia Beach, VA (3% average, 7.5% maximum) have the highest potential revenue loss from the Active Projects based on 2008-2024 total port fisheries revenue. In response to Council request, we also [present information](https://noaa-edab.github.io/catalog/wind_port.html) on the New England ports that land MAFMC-managed species within the wind lease areas. This information can help fisheries managers better understand how MAFMC-managed species may be more broadly affected by offshore wind development. Seven New England ports attribute the majority of their landings within the wind lease areas to MAFMC-managed species (Fig. \ref{fig:wind-rev}); note that two of these ports do not have a majority of MAFMC-managed landings when only the Active Project areas are considered. Additionally, the Active Projects account for less than 8% of total revenue in these seven ports. Future analyses should also consider that the impacts of offshore wind development may unevenly affect individual operators, with some permit holders deriving a much higher proportion of revenue from wind areas than the port-based mean.

<!-- \newpage -->

```{r wea-landings-rev, ft.arraystretch = 1}
spp.mark = c('Chub Mackerel','Bluefish','Golden Tilefish')

create_wea_landings_rev_table(council = "MAFMC") |>
  dplyr::arrange(desc(`Maximum Percent Total Annual Regional Species Revenue`)) |> 
  dplyr::slice_head(n = 10) |>
  dplyr::mutate(
    `NEFMC, MAFMC, and ASMFC Managed Species` = dplyr::case_when(
      `NEFMC, MAFMC, and ASMFC Managed Species` %in% spp.mark ~ paste0(`NEFMC, MAFMC, and ASMFC Managed Species`, "*"),
      TRUE ~ `NEFMC, MAFMC, and ASMFC Managed Species`
    )
  ) |>
  flextable::flextable() %>%
  flextable::set_caption("Top 10 Mid-Atlantic managed species landings and revenue from Active Projects. *These species did not have more than 50,000 lbs landed from the Active Project areas in any year.") %>%
  flextable::width(width = c(2,2,2))
```

#### Coastal community vulnerability

`r socio_intro` Point Lookout, NY and Newport News, VA have [socio-demographic or gentrification concerns](https://noaa-edab.github.io/catalog/engagement.html) and should be recognized as having additional vulnerability from Active Projects. If all proposed projects are developed, BOEM reports that cumulative offshore wind development could have moderate impacts on low-income members of communities who work in the commercial fishing and for-hire fishing industry due to disruptions to fish populations, restrictions on navigation and increased vessel traffic, and exacerbation of existing vulnerabilities of low-income workers to economic impacts. Historically, the introduction of new industries can trigger industrial and socioeconomic gentrification of fishing ports. Some of the communities with the highest fisheries revenue overlap with offshore wind development areas that are also vulnerable to gentrification pressure are `r gentrification_ports`. Competition for port space and potential pivoting of space use for offshore wind development should be monitored closely to ensure fishing communities are not adversely impacted. 

Additionally, offshore wind could increase recreational fishing opportunities at the turbines, potentially creating a demand for additional tourism, recreational fishing and boating port space in communities already balancing these uses with commercial fishing infrastructure (e.g., Virginia Beach, VA, Montauk, NY and Barnegat Light, NJ). 

```{r wea-port-rev, fig.cap=return_caption(chunk_name = "wea-port-rev", region = "MidAtlantic"),  fig.width=6.5, fig.asp=.82, fig.align='left'}
return_plot("wea-port-rev", region = "MidAtlantic")
```

```{r wind-rev, fig.cap=return_caption(chunk_name = "wind-rev", region = "MidAtlantic"), fig.width=6.5, fig.asp=.35, fig.align='left'}
return_plot("wind-rev", region = "MidAtlantic")
```

```{r, eval = params$id_child_docs, results = "asis"}
<<child-footer>>
```
